---
title: "Exploratory Plots"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    fig_caption: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
  # bookdown::pdf_document2:
  #   highlight: rstudio
---

# References


# TODO

- Load data +
- Reformat +
- Select dances of interest +
- Plot dances of interest +
- Feeder angle plotting function
- Solar ephemeris plots
- Elevation-accuracy plots

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load functions and packages

Load the packages and other functions that will be used during reorganisation.

```{r load-functions}
source('HS_functions.R')
```


# Load data


## Load the raw data

Load the raw data exported from Excel to `csv`.

```{r load-the-data}
path_file = "0Data/hs_dance_reorg.csv"
if(!file.exists(path_file))
{print(path_file, '\nnot found')}
hd = read.table(file = path_file, 
                header = T, 
                sep  = ',')
```


## Inspect data

Check the names of variables in this dataset.

```{r check-names}
names(hd)
```

<details>

<summary>Click to expand raw data</summary>

```{r inspect-data}
print(hd)
```


# Format data

Ensure that each variable has the correct data format for subsequent analysis.

```{r Data formatting}
hd = within(hd,
            {
              ID = as.factor(ID) # bee identifier as a factor
              ID_d = as.factor(ID_d) # beedance identifier as a factor
              day = as.factor(day) # date as a factor
              sast_time = as.POSIXct(sast_time,
                                     tz = "Africa/Johannesburg", 
                                      format = "%Y-%m-%d %H:%M:%OS")#South African Standard time
            }
)
u_id = with(hd, unique(ID)) #unique dancers
length(u_id)#40 bees
u_id_d = with(hd, unique(ID_d)) # unique beedances
length(u_id_d)#377 beedances
```

# Plot Examples

## Set up plot parameters

In each plot, we will indicate the condition and time of day using different colours. Experiment times will be added for reference (in seconds after midnight).
```{r plot-parameters}
#condition colours
cond_cols = list(
  vertical = 'darkred',
  horizontal = 'darkgreen',
  tilted = 'seagreen',
  solar = 'blue4',
  antisolar = 'cyan4',
  zenith = "skyblue3"
)


#assign experiment times
morning_start = 7.50 * 60 * 60
morning_end = 9.00 * 60 * 60
midday_start = 11.00 * 60 * 60
midday_end = 13.00 * 60 * 60
afternoon_start = 15.00 * 60 * 60
afternoon_end = 16.50 * 60 * 60
evening_start = 17.50 * 60 * 60
evening_end = 19.00 * 60 * 60

exper_times = c(
  morning_start = morning_start,
  morning_end = morning_end,
  midday_start = midday_start,
  midday_end = midday_end,
  afternoon_start = afternoon_start,
  afternoon_end = afternoon_end,
  evening_start = evening_start,
  evening_end = evening_end
)


time_cols = c(
  morning_start   = "#C68B59",
  morning_end     = "#C68B59",
  midday_start    = "#E6C594",
  midday_end      = "#E6C594",
  afternoon_start = "#C68B59",
  afternoon_end   = "#C68B59",
  evening_start   = "#94B49F",
  evening_end     = "#94B49F"
)


```


## Plot an example dance

For example we can look at the dances for Blue-22 recorded on 2025-01-28 at 
08:12, 08:36, 08:58, 08:34 & 11:44—each of which is the first for one of the conditions.
Dances are plotted relative to sun position (0°) and feeder azimuth is shown as an brown line.

```{r example-subset}
example_day  = unique(hd$day)[length(unique(hd$day))]#last day
example_bee = "22b"
print(paste("First dance for", example_bee, "for each condition on", example_day))
daybee_lst = FirstDanceCond(dta = hd,
                        subset = day %in% example_day & ID %in% example_bee)  
# Plot the example subset
Plt_subs(subs = daybee_lst$example_subs,
         angle_for = "sun",
         cond_cols = cond_cols,
         title = paste("First dance for", example_bee, "for each condition on", example_day)
         )
```
At least for this subset, the dances follow the expected pattern.
For both vertical and horizontal dances, the average waggle run aligns with 
the feeder direction. With antisolar skylight the waggle runs are also aligned,
while solar skylight is mistaken for antisolar skylight, resulting in a 180° turn.


We can also plot these dances with the feeder direction subtracted,
so that the feeder falls at 0°.

```{r feeder-angle-examples}
#Plot using the feeder as the frame of reference
Plt_subs(subs = daybee_lst$example_subs,
         angle_for = "feeder",
         title = paste("Feeder angles for 1st dance of", example_bee, "for each condition on", example_day))
```


## Low elevation


```{r low-elevation-examples}
# Select the first dance for each condition at low sun elevation.
print("First dance for low sun elevations")
lowel_subs = FirstDanceCond(
  hd,
  sun_el < 15
)$example_subs

Plt_subs(
  subs = lowel_subs,
  angle_for = "feeder",
  title = paste("Feeder angles for low sun elevations")
)
```



## Medium elevation


```{r low-elevation-examples}
# Select the first dance for each condition at medium sun elevation.
print(paste("First dance for medium sun elevations"))
medel_subs = FirstDanceCond(
  hd,
  sun_el > 30 & sun_el < 45
)$example_subs

Plt_subs(
  subs = medel_subs,
  angle_for = "feeder",
  title = paste("Feeder angles for medium sun elevations")
)
```




## High elevation


```{r low-elevation-examples}
# Select the first dance for each condition at high sun elevation.
print(paste("First dance for high sun elevations"))
higel_subs = FirstDanceCond(
  hd,
  sun_el > 75
)$example_subs

Plt_subs(
  subs = higel_subs,
  angle_for = "feeder",
  title = paste("Feeder angles for high sun elevations")
)
```




# Solar Ephemeris

Using the vertical dances, it should be possible to identify the solar ephemeris 
functions used by those individuals that danced repeatedly throughout the day.


## Check dance times
Blue 22 and blue 66 both danced very frequently, so we can start by checking when they danced.


```{r vertical-dance-times}

par(mfrow = c(1,1))
#plot all vertical dances
with(subset(hd, condition %in% c("vertical") ),
     {
     plot(x = GetToD(sast_time,
                     tz = 'UTC'),
          y = Mod360.180(sun_az),
          xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
          ylim = c(-1,1)*180,
          xlab = 'UTC',
          ylab = 'Sun Azimuth',
          main = 'Vertical dances',
          pch = 3,#cross
          col = 'gray50'
     )
     abline(h = -4:4*c(45),
            lty = 2) #limits
     abline(v =  12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
            lty = 3)
     }
)
#add lines indicating the approximate expected start and end of each experiment
abline(v = exper_times,
       col = time_cols,
       lty = c(1,2),
       lwd = 2
       )
#plot dances by bees of interest
with(subset(hd, condition %in% c("vertical") & 
              ID %in% c('22b', '66b') 
            ),
     {
     points(x = GetToD(sast_time,
                     tz = 'UTC'),
          y = Mod360.180(sun_az),
          pch = 21,
          col = 'black',
          cex = 1.2,
           bg = adjustcolor(sapply(X = as.character(ID),
                        FUN = switch,
                        `22b` = 'cyan',
                        `66b` = 'cadetblue',
                        'red',#if these show up, oops!
                        alpha.f = 0.25)
           )
     )
     }
)
legend(x = 'topright',
       legend = c("Blue 22",
                  "Blue 66",
                  "other"),
       col=  c("cyan",
               "cadetblue",
               "gray50"),
       pch = c(19, 19,
               3)
)


```


It looks like we have the best chance with blue 22, for which we have 
the most dances during the rapid azimuth change period in the middle of the day.


```{r b22-ephemeris}

par(mfrow = c(1,1))
#plot all vertical dances
with(hd,
     {
     plot(x = sort(GetToD(sast_time,
                     tz = 'UTC')),
          y = Mod360.180(sun_az)[order(GetToD(sast_time,
                     tz = 'UTC'))],
          xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
          ylim = c(-1,1)*180,
          xlab = 'UTC',
          ylab = 'Sun Azimuth',
          main = 'Blue 22, Vertical Dances',
          type = 'b',
          pch = 3,#cross
          lwd = 5,
          col = 'gray50'
     )
     abline(h = -4:4*c(45),
            lty = 2) #limits
     abline(v =  12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
            lty = 3)
     }
)
#add lines indicating the approximate expected start and end of each experiment
abline(v = exper_times,
       col = time_cols,
       lty = c(1,2),
       lwd = 2
       )
#plot dances by bees of interest
with(subset(hd, condition %in% c("vertical") &
              ID %in% c('22b')
            ),
     {
     points(x = GetToD(sast_time,
                     tz = 'UTC'),
          y = Mod360.180(dance_sun_angle + (sun_az - feeder_az)),
          pch = 21,
          cex = 1.2,
          col = adjustcolor('black',
                            alpha.f = 0.5),
           bg = adjustcolor('cyan',
                        alpha.f = 0.25)
     )
     }
)
legend(
  x = 'bottomright',
  legend = c(
    "Sun azimuth",
    "Implied azimuth"
  ),
  col = c("gray50", "cyan"),
  pch = c(3, 19),
  cex = 0.7
)

```
