---
title: "Exploratory Plots"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    fig_caption: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
  # bookdown::pdf_document2:
  #   highlight: rstudio
---

# References


# TODO

- Load data +
- Reformat +
- Select dances of interest +
- Plot dances of interest +
- Feeder angle plotting function +
- Solar ephemeris plots +
- Elevation-accuracy plots +
- Add bimodality 
- Paired dances
- All dances

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load functions and packages

Load the packages and other functions that will be used during reorganisation.

```{r load-functions}
source('HS_functions.R')
```


# Load data


## Load the raw data

Load the raw data exported from Excel to `csv`.

```{r load-the-data}
path_file = "0Data/hs_dance_reorg.csv"
if(!file.exists(path_file))
{print(path_file, '\nnot found')}
hd = read.table(file = path_file, 
                header = T, 
                sep  = ',')
```


## Inspect data

Check the names of variables in this dataset.

```{r check-names}
names(hd)
```

<details>

<summary>Click to expand raw data</summary>

```{r inspect-data}
print(hd)
```


# Format data

Ensure that each variable has the correct data format for subsequent analysis.

```{r Data formatting}
hd = within(hd, {
  ID = as.factor(ID) # bee identifier as a factor
  ID_d = as.factor(ID_d) # beedance identifier as a factor
  day = as.factor(day) # date as a factor
  condition = factor(
    condition,
    levels = c(
      "vertical",
      "horizontal",
      "tilted",
      "solar",
      "antisolar",
      "zenith"
    )
  ) #set up conditions as a factor in preferred order
  sast_time = as.POSIXct(
    sast_time,
    tz = "Africa/Johannesburg",
    format = "%Y-%m-%d %H:%M:%OS"
  ) #South African Standard time
  cangle = circular(
    feeder_angle,
    template = 'none'
  ) #circular formatted angle for circular functions
})

u_id = with(hd, unique(ID)) #unique dancers
length(u_id)#40 bees
u_id_d = with(hd, unique(ID_d)) # unique beedances
length(u_id_d)#377 beedances
```

# Plot Examples

## Set up plot parameters

In each plot, we will indicate the condition and time of day using different colours. Experiment times will be added for reference (in seconds after midnight).
```{r plot-parameters}
#condition colours
cond_cols = list(
  vertical = 'darkred',
  horizontal = 'darkgreen',
  tilted = 'seagreen',
  solar = 'blue4',
  antisolar = 'cyan4',
  zenith = "skyblue3"
)


#assign experiment times
morning_start = 7.50 * 60 * 60
morning_end = 9.00 * 60 * 60
midday_start = 11.00 * 60 * 60
midday_end = 13.00 * 60 * 60
afternoon_start = 15.00 * 60 * 60
afternoon_end = 16.50 * 60 * 60
evening_start = 17.50 * 60 * 60
evening_end = 19.00 * 60 * 60

exper_times = c(
  morning_start = morning_start,
  morning_end = morning_end,
  midday_start = midday_start,
  midday_end = midday_end,
  afternoon_start = afternoon_start,
  afternoon_end = afternoon_end,
  evening_start = evening_start,
  evening_end = evening_end
)


time_cols = c(
  morning_start   = "#C68B59",
  morning_end     = "#C68B59",
  midday_start    = "#E6C594",
  midday_end      = "#E6C594",
  afternoon_start = "#C68B59",
  afternoon_end   = "#C68B59",
  evening_start   = "#94B49F",
  evening_end     = "#94B49F"
)


```


## Plot an example dance

For example we can look at the dances for Blue-22 recorded on 2025-01-28 at 
08:12, 08:36, 08:58, 08:34 & 11:44—each of which is the first for one of the conditions.
Dances are plotted relative to sun position (0°) and feeder azimuth is shown as an brown line.

```{r example-subset}
example_day  = unique(hd$day)[length(unique(hd$day))]#last day
example_bee = "22b"
print(paste("First dance for", example_bee, "for each condition on", example_day))
daybee_lst = FirstDanceCond(dta = hd,
                        subset = day %in% example_day & ID %in% example_bee)  
# Plot the example subset
Plt_subs(subs = daybee_lst$example_subs,
         angle_for = "sun",
         cond_cols = cond_cols,
         title = paste("First dance for", example_bee, "for each condition on", example_day)
         )
```
At least for this subset, the dances follow the expected pattern.
For both vertical and horizontal dances, the average waggle run aligns with 
the feeder direction. With antisolar skylight the waggle runs are also aligned,
while solar skylight is mistaken for antisolar skylight, resulting in a 180° turn.


We can also plot these dances with the feeder direction subtracted,
so that the feeder falls at 0°.

```{r feeder-angle-examples}
#Plot using the feeder as the frame of reference
Plt_subs(subs = daybee_lst$example_subs,
         angle_for = "feeder",
         title = paste("Feeder angles for 1st dance of", example_bee, "for each condition on", example_day))
```


## Low elevation


```{r low-elevation-examples}
# Select the first dance for each condition at low sun elevation.
print("First dance for low sun elevations")
lowel_subs = FirstDanceCond(
  hd,
  sun_el < 15 
)$example_subs

Plt_subs(
  subs = lowel_subs,
  angle_for = "feeder",
  title = paste("Feeder angles for low sun elevations")
)
```



## Medium elevation


```{r medium-elevation-examples}
# Select the first dance for each condition at medium sun elevation.
print(paste("First dance for medium sun elevations"))
medel_subs = FirstDanceCond(
  hd,
  sun_el > 30 & sun_el < 45
)$example_subs

Plt_subs(
  subs = medel_subs,
  angle_for = "feeder",
  title = paste("Feeder angles for medium sun elevations")
)
```




## High elevation


```{r high-elevation-examples}
# Select the first dance for each condition at high sun elevation.
print(paste("First dance for high sun elevations"))
higel_subs = FirstDanceCond(
  hd,
  sun_el > 75
)$example_subs

Plt_subs(
  subs = higel_subs,
  angle_for = "feeder",
  title = paste("Feeder angles for high sun elevations")
)
```


# Plot the raw dance data

## Plot an example dance

For example we can look at the longest dance in the dataset,  by Blue 73, 
recorded on 2025.01.19 at 16:51.

```{r, example-dance}
longest_id_d = u_id_d[which.max(sapply(
  u_id_d,
  function(i) {
    with(
      subset(hd, ID_d %in% i),
      max(waggle_run)
    )
  }
))]

par(pty = 's')
par(mfrow = c(1,1), mar = c(0,0,0,0))
PltDance(ids = '73b_00027_16:51:00_20250119',
         dtf = hd,
         rlim = 100,
         cx = 1.0)
segments(x0 = 0, x1 = 0,
        y0 = 0, y1 = 100,
        col = "yellow3")
with(subset(hd, ID_d %in% '73b_00027_16:51:00_20250119' & waggle_run %in% c(20, 40, 60, 80)),
     {
points(x = waggle_run*sin(rad(dance_sun_angle)),
     y = waggle_run*cos(rad(dance_sun_angle)),
     labels = paste(' run', waggle_run),
     col = hcl.colors(4, palette = "dark2"),
     pch = 19)
text(x = waggle_run*sin(rad(dance_sun_angle)),
     y = waggle_run*cos(rad(dance_sun_angle)),
     labels = paste(' run', waggle_run),
     col = hcl.colors(4, palette = "dark2"),
     cex = 0.8,
     adj = c(0,0))
     }
)
legend(x = 'topright',
       legend = c('Waggle run',
                  'Mean dance direction',
                  'Geographic North',
                  'Sun azimuth (0°)',
                  'Feeder direction'),
       col = c('darkred',
               'darkred',
               'black',
               'yellow3',
               'navajowhite4'
               ),
       pch = c(21,NA,NA,NA,NA),
       lty = c(NA, 1, 1, 1, 1),
       lwd = c(2,3,3,3,3),
       cex = 0.5
       )
```

## Plot all raw dances

<details>

<summary>Click to expand dance plots</summary>


```{r Raw dances, warning=FALSE}
#reorganise u_id_d by date of dance
dance_date = sapply(
  X = u_id_d,
  FUN = function(i) {
    with(
      subset(hd, ID_d %in% i),
      unique(sast_time)
    )
  }
)
u_id_d = u_id_d[order(dance_date)]

par(mfrow = c(2, 2),
    mar = c(0,0,0,0))
par(pty = 's')
#add legend to 1st page
plot(x = NULL,
     xlim = c(-1,1),
     ylim = c(-1,1),
     pch = 19,
     axes = FALSE,
     xlab = '',
     ylab = '',
     main = ''
)
legend(x = 'center',
       legend = c('Waggle run',
                  'Mean dance direction',
                  'Geographic North',
                  'Sun azimuth (0°)',
                  'Feeder direction'),
       col = c('darkred',
               'darkred',
               'black',
               'yellow3',
               'navajowhite4'
               ),
       pch = c(21,NA,NA,NA,NA),
       lty = c(NA, 1, 1, 1, 1),
       lwd = c(2,3,3,3,3),
       cex = 1.3
       )
#loop through individuals
PltDance(ids = u_id_d,
         dtf = hd,
         cx = 1.0)
#save as a plot
invisible({
  pdf(
    file = "Plots/All_dances.pdf",
    paper = 'a4',
    onefile = TRUE
  )
  pdfdev = dev.cur()
  par(
    mfrow = c(2, 2),
    mar = c(0, 0, 0, 0)
  )
  par(pty = 's')

  #add legend to 1st page
  plot(
    x = NULL,
    xlim = c(-1, 1),
    ylim = c(-1, 1),
    pch = 19,
    axes = FALSE,
    xlab = '',
    ylab = '',
    main = ''
  )
  legend(
    x = 'center',
    legend = c(
      'Waggle run',
      'Mean dance direction',
      'Geographic North',
      'Sun azimuth (0°)',
      'Feeder direction'
    ),
    col = c(
      'darkred',
      'darkred',
      'black',
      'yellow3',
      'navajowhite4'
    ),
    pch = c(21, NA, NA, NA, NA),
    lty = c(NA, 1, 1, 1, 1),
    lwd = c(2, 3, 3, 3, 3),
    cex = 2.0
  )
  #loop through individuals
  PltDance(
    ids = u_id_d,
    dtf = hd,
    cx = 1.0
  )
  dev.off(which = pdfdev)
})
         
```


# Solar Ephemeris

Using the vertical dances, it should be possible to identify the solar ephemeris 
functions used by those individuals that danced repeatedly throughout the day.

First set up the reference solar ephemeris for all times on a median day.

```{r generate-ephemeris}
hive = r"(25° 34' 18" S, 31° 10' 48" E)" # N.B. this string contains literal " so needs wrapping in r"()"

# Define coordinates (lon, lat)
hive = Dms2decimal(hive) 


# sequence from times from 1h before 1st experiment to 1h after last, by 5 minutes
time_seq = seq(from = min(exper_times)-60*60,
               to  = max(exper_times)+60*60,
               by = 5*60)
med_date = with(hd, median(as.numeric(paste(day))))
time_code_sast = as.POSIXct(paste(med_date, "00:00:00"),#midnight on median day 
                                      tz = "Africa/Johannesburg", 
                                      format = "%Y%m%d %H:%M:%OS") + #South African Standard time
              time_seq
#convert to UTC
time_code_utc = as.POSIXct(time_code_sast, tz = "UTC")#UTC time

saz_seq = mapply(
                   FUN = GetSaz,
                   tm = time_code_utc,
                   lon = hive['longitude'],
                   lat = hive['latitude']
                   )
par(mfrow = c(1,1))
#plot all vertical dances

plot(x = GetToD(format(time_code_sast, "%H:%M:%S")),
    y = Mod360.180(saz_seq),
    xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
    ylim = c(-1,1)*180,
    xlab = 'Time of Day',
    ylab = 'Sun Azimuth',
    main = 'Median Ephemeris',
    type = 'l',
    col = 'yellow3',
    lwd = 5
)
abline(h = -4:4*c(45),
      lty = 2) #limits
abline(v =  12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
      lty = 3)

#add lines indicating the approximate expected start and end of each experiment
abline(v = exper_times,
       col = time_cols,
       lty = c(1,2),
       lwd = 2
       )


```


## Check dance times
Blue 22 and blue 66 both danced very frequently, so we can start by checking when they danced.


```{r vertical-dance-times}


par(mfrow = c(1,1))
#plot all vertical dances
plot(x = GetToD(format(time_code_sast, "%H:%M:%S")),
    y = Mod360.180(saz_seq),
    xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
    ylim = c(-1,1)*180,
    xlab = 'Time of Day',
    ylab = 'Sun Azimuth',
    main = 'Vertical dances',
    type = 'l',
    col = 'yellow3',
    lwd = 7
)

with(subset(hd, condition %in% c("vertical") ),
     {
     points(x = GetToD(time),
          y = Mod360.180(sun_az),
          pch = 3,#cross
          col = 'gray25'
     )
     abline(h = -4:4*c(45),
            lty = 2) #limits
     abline(v =  12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
            lty = 3)
     }
)
#add lines indicating the approximate expected start and end of each experiment
abline(v = exper_times,
       col = time_cols,
       lty = c(1,2),
       lwd = 2
       )
#plot dances by bees of interest
with(subset(hd, condition %in% c("vertical") & 
              ID %in% c('22b', '66b') 
            ),
     {
     points(x = GetToD(time),
          y = Mod360.180(sun_az),
          pch = 21,
          col = 'black',
          cex = 1.2,
           bg = adjustcolor(sapply(X = as.character(ID),
                        FUN = switch,
                        `22b` = 'cyan',
                        `66b` = 'cadetblue',
                        'red',#if these show up, oops!
                        alpha.f = 0.25)
           )
     )
     }
)
legend(x = 'topright',
       legend = c("Blue 22",
                  "Blue 66",
                  "other"),
       col=  c("cyan",
               "cadetblue",
               "gray25"),
       pch = c(19, 19,
               3)
)


```


It looks like we have the best chance with blue 22, for which we have 
the most dances during the rapid azimuth change period in the middle of the day.


```{r b22-ephemeris}

par(mfrow = c(1,1))
#plot all vertical dances
plot(x = GetToD(format(time_code_sast, "%H:%M:%S")),
    y = Mod360.180(saz_seq),
    xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
    ylim = c(-1,1)*180,
    xlab = 'Time of Day',
    ylab = 'Sun Azimuth',
    main = 'Blue 22, Vertical Dances',
    type = 'l',
    col = 'yellow3',
    lwd = 7
)
with(subset(hd, condition %in% c("vertical") &
              ID %in% c('22b')
            ),
     {
     points(x = sort(GetToD(time)),
          y = Mod360.180(sun_az)[order(GetToD(time))],
          pch = 3,#cross
          lwd = 3,
          col = 'gray25'
     )
     abline(h = -4:4*c(45),
            lty = 2) #limits
     abline(v =  12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
            lty = 3)
     }
)
#add lines indicating the approximate expected start and end of each experiment
abline(v = exper_times,
       col = time_cols,
       lty = c(1,2),
       lwd = 2
       )
#plot dances by bees of interest
with(subset(hd, condition %in% c("vertical") &
              ID %in% c('22b')
            ),
     {
     points(x = GetToD(time),
          y = Mod360.180(dance_sun_angle + (sun_az - feeder_az)),
          pch = 21,
          cex = 1.2,
          col = adjustcolor('black',
                            alpha.f = 0.5),
           bg = adjustcolor('cyan',
                        alpha.f = 0.25)
     )
     }
)
legend(
  x = 'bottomright',
  legend = c(
    "Sun azimuth",
    "Implied azimuth"
  ),
  col = c("gray25", "cyan"),
  pch = c(3, 19),
  cex = 0.7
)

```

How similar is the population?

```{r all-ephemeris}

par(mfrow = c(1,1))
#plot all vertical dances
plot(x = GetToD(format(time_code_sast, "%H:%M:%S")),
    y = Mod360.180(saz_seq),
    xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
    ylim = c(-1,1)*180,
    xlab = 'Time of Day',
    ylab = 'Sun Azimuth',
    main = 'All Bees, Vertical Dances',
    type = 'l',
    col = 'yellow3',
    lwd = 7
)
with(hd,
     {
     points(x = sort(GetToD(time)),
          y = Mod360.180(sun_az)[order(GetToD(time))],
          pch = 3,#cross
          lwd = 3,
          col = 'gray25'
     )
     abline(h = -4:4*c(45),
            lty = 2) #limits
     abline(v =  12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
            lty = 3)
     }
)
#add lines indicating the approximate expected start and end of each experiment
abline(v = exper_times,
       col = time_cols,
       lty = c(1,2),
       lwd = 2
       )
#plot dances other bees
with(subset(hd, condition %in% c("vertical") &
            !(ID %in% c('22b')) ),
     {
     points(x = GetToD(time),
          y = Mod360.180(dance_sun_angle + (sun_az - feeder_az)),
          pch = 21,
          cex = 1.2,
          col = adjustcolor('black',
                            alpha.f = 0.1),
           bg = adjustcolor('cadetblue',
                        alpha.f = 0.1)
     )
     }
)

#plot dances by bees of interest
with(subset(hd, condition %in% c("vertical") &
              ID %in% c('22b')
            ),
     {
     points(x = GetToD(time),
          y = Mod360.180(dance_sun_angle + (sun_az - feeder_az)),
          pch = 21,
          cex = 0.5,
          col = adjustcolor('black',
                            alpha.f = 0.5),
           bg = adjustcolor('cyan',
                        alpha.f = 1.0)
     )
     }
)

legend(
  x = 'bottomright',
  legend = c(
    "Sun azimuth",
    "B22",
    "Other bees"
  ),
  col = c("gray25", "cyan", "cadetblue"),
  pch = c(3, 19, 19),
  cex = 0.7
)

```


It looks like B22 is fairly representative:
all bees are initially slow to adjust to the sun's accelerating westward movement 
after 11:00, but have adjusted by 12:00. 
In the afternoon dances start at the correct angle, appear not to adjust at all 
during the following hour, but begin to adjust around 17:30.
A simple description could be achieved with a piecewise regression, 
with a constant rate of change in solar azimuth between 07:00 and 11:00, 
another (faster) rate of change between 11:00 and 13:00, 
and likely the same morning rate of change repeated again between 13:00 and 19:00.

## Ephemeris error

What does the error distribution look like? 
Is it symmetrical, or is there a systematic offset?


```{r calculate-ephemeris-error, fig.width = 5, fig.height = 7, dpi = 150}

par(
  mfrow = c(2, 1),
  mar = c(4, 4, 0, 0)
)
#plot dance offset from true feeder angle
with(
  subset(
    hd,
    condition %in% c("vertical")
  ),
  {
    plot(
      x = GetToD(time),
      y = Mod360.180(deg(feeder_angle)),
      xlim = c(20400, 78060), #"06:40:00 CET" "22:41:00 CET"
      ylim = c(-1, 1) * 180,
      xlab = 'Time of Day',
      ylab = 'Waggle run - sun azimuth',
      main = '', #'Vertical Dances',
      pch = 21,
      cex = 1.2,
      col = adjustcolor(
        'black',
        alpha.f = 0.1
      ),
      bg = adjustcolor(
        'cadetblue',
        alpha.f = 0.1
      )
    )
  }
)
abline(h = 0, lwd = 3) #expected
abline(h = -4:4 * c(45), lty = 2) #limits
abline(v = 7:19 * 60*60, lty = 3) #hours
abline(
  v = 12 * 60 * 60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
  lty = 3
)
with(
  subset(
    hd,
    condition %in% c("vertical")
  ),
  {
    lines(
      loess.smooth(
        x = GetToD(
          time
        ),
        y = Mod360.180(deg(
          feeder_angle
        ))
      ),

      col = 'darkblue',
      lwd = 5
    )
  }
)

#absolute error

#plot dances other bees
with(
  subset(
    hd,
    condition %in% c("vertical")
  ),
  {
    plot(
      x = GetToD(time),
      y = abs(Mod360.180(deg(
        feeder_angle
      ))),
      xlim = c(20400, 78060), #"06:40:00 CET" "22:41:00 CET"
      ylim = c(0, 1) * 180,
      xlab = 'Time of Day',
      ylab = 'absolute feeder angle error',
      main = '', #'Vertical Dances',
      pch = 21,
      cex = 1.2,
      col = adjustcolor(
        'black',
        alpha.f = 0.1
      ),
      bg = adjustcolor(
        'cadetblue',
        alpha.f = 0.1
      )
    )
  }
)
abline(h = 0, lwd = 3) #expected
abline(h = 2:12 * c(15), lty = 2) #limits
abline(v = 7:19 * 60*60, lty = 3) #hours
abline(
  v = 12 * 60 * 60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
  lty = 3
)
with(
  subset(
    hd,
    condition %in% c("vertical")
  ),
  {
    lines(
      LogitLoess(
        x = GetToD(
          time
        ),
        y = abs(Mod360.180(deg(
          feeder_angle
        ))),
        absangle = TRUE
      ),

      col = 'darkblue',
      lwd = 5
    )
  }
)

```

## Horizontal dances

Is this similar for the horizontal

```{r all-ephemeris-horizontal}

par(mfrow = c(1,1))
#plot all vertical dances
plot(x = GetToD(format(time_code_sast, "%H:%M:%S")),
    y = Mod360.180(saz_seq),
    xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
    ylim = c(-1,1)*180,
    xlab = 'Time of Day',
    ylab = 'Sun Azimuth',
    main = 'All Bees, Horizontal Dances',
    type = 'l',
    col = 'yellow3',
    lwd = 7
)
with(hd,
     {
     points(x = sort(GetToD(time)),
          y = Mod360.180(sun_az)[order(GetToD(time))],
          pch = 3,#cross
          lwd = 3,
          col = 'gray25'
     )
     abline(h = -4:4*c(45),
            lty = 2) #limits
     abline(v =  12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
            lty = 3)
     }
)
#add lines indicating the approximate expected start and end of each experiment
abline(v = exper_times,
       col = time_cols,
       lty = c(1,2),
       lwd = 2
       )
#plot dances other bees
with(subset(hd, condition %in% c("horizontal") &
            !(ID %in% c('22b')) ),
     {
     points(x = GetToD(time),
          y = Mod360.180(dance_sun_angle + (sun_az - feeder_az)),
          pch = 21,
          cex = 1.2,
          col = adjustcolor('black',
                            alpha.f = 0.1),
           bg = adjustcolor('cadetblue',
                        alpha.f = 0.1)
     )
     }
)

#plot dances by bees of interest
with(subset(hd, condition %in% c("horizontal") &
              ID %in% c('22b')
            ),
     {
     points(x = GetToD(time),
          y = Mod360.180(dance_sun_angle + (sun_az - feeder_az)),
          pch = 21,
          cex = 0.5,
          col = adjustcolor('black',
                            alpha.f = 0.5),
           bg = adjustcolor('cyan',
                        alpha.f = 1.0)
     )
     }
)

legend(
  x = 'bottomright',
  legend = c(
    "Sun azimuth",
    "B22",
    "Other bees"
  ),
  col = c("gray25", "cyan", "cadetblue"),
  pch = c(3, 19, 19),
  cex = 0.7
)
```

```{r calculate-ephemeris-error-horizontal, fig.width = 5, fig.height = 7, dpi = 150}

par(
  mfrow = c(2, 1),
  mar = c(4, 4, 2, 0)
)
#plot dance offset from true feeder angle
with(
  subset(
    hd,
    condition %in% c("horizontal")
  ),
  {
    plot(
      x = GetToD(time),
      y = Mod360.180(deg(feeder_angle)),
      xlim = c(20400, 78060), #"06:40:00 CET" "22:41:00 CET"
      ylim = c(-1, 1) * 180,
      xlab = 'Time of Day',
      ylab = 'Waggle run - feeder azimuth',
      main = 'Horizontal Dances',
      pch = 21,
      cex = 1.2,
      col = adjustcolor(
        'black',
        alpha.f = 0.1
      ),
      bg = adjustcolor(
        'cadetblue',
        alpha.f = 0.1
      )
    )
  }
)
abline(h = 0, lwd = 3) #expected
abline(h = -4:4 * c(45), lty = 2) #limits
abline(v = 7:19 * 60*60, lty = 3) #hours
abline(
  v = 12 * 60 * 60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
  lty = 3
)
with(
  subset(
    hd,
    condition %in% c("horizontal")
  ),
  {
    lines(
      loess.smooth(
        x = GetToD(
          time
        ),
        y = Mod360.180(deg(
          feeder_angle
        ))
      ),

      col = 'darkblue',
      lwd = 5
    )
  }
)

#absolute error

#plot dances other bees
with(
  subset(
    hd,
    condition %in% c("horizontal")
  ),
  {
    plot(
      x = GetToD(time),
      y = abs(Mod360.180(deg(
        feeder_angle
      ))),
      xlim = c(20400, 78060), #"06:40:00 CET" "22:41:00 CET"
      ylim = c(0, 1) * 180,
      xlab = 'Time of Day',
      ylab = 'absolute feeder angle error',
      main = 'Horizontal Dances',
      pch = 21,
      cex = 1.2,
      col = adjustcolor(
        'black',
        alpha.f = 0.1
      ),
      bg = adjustcolor(
        'cadetblue',
        alpha.f = 0.1
      )
    )
  }
)
abline(h = 0, lwd = 3) #expected
abline(h = 2:12 * c(15), lty = 2) #limits
abline(v = 7:19 * 60*60, lty = 3) #hours
abline(
  v = 12 * 60 * 60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
  lty = 3
)
with(
  subset(
    hd,
    condition %in% c("horizontal")
  ),
  {
    lines(
      LogitLoess(
        x = GetToD(
          time
        ),
        y = abs(Mod360.180(deg(
          feeder_angle
        ))),
        absangle = TRUE
      ),

      col = 'darkblue',
      lwd = 5
    )
  }
)

```



# Calculate dance means


```{r dance-mean-vectors}
#calculate mean vectors
mean_vectors = aggregate(cangle~ID_d+ #for each dance 
                           ID+condition+sun_az+sun_el + # also save these variables
                           time+day,
                         data = hd,
                         FUN = rho.circular
)
#correct names
mean_vectors = within(mean_vectors,
                      {mean_vector = cangle; rm(cangle)} # angle now indicates a mean vector, not an angle
)
#add kappa
mle_estimates = aggregate(cangle~ID_d+ #for each dance 
                           ID+condition+sun_az+sun_el + # also save these variables
                           time+day,
                          data = hd,
                          FUN = MLE_est
)
#Rayleigh tests
rayleigh_tests = aggregate(cangle~ID_d+ #for each dance 
                           ID+condition+sun_az+sun_el + # also save these variables
                           time+day,
                          data = hd,
                          FUN = Rayleigh_p
                          )

#correct names
rayleigh_tests = within(rayleigh_tests,
                      {p_rayleigh = cangle; rm(cangle)} # angle now indicates a mean vector, not an angle
)

#add to the summary table and 
#calculate inverse softplus kappa
mean_vectors = within(mean_vectors,
                      {
                        mu = deg(mle_estimates$cangle[,'mu'])
                        kappa = mle_estimates$cangle[,'kappa']
                      }
)
#check how many dances would count as "oriented"
print(paste("Rayleigh p<0.05, proportion of all dances:") )
#all dances
with(rayleigh_tests,
     print( mean(p_rayleigh <0.05) *100,digits = 2)
     )

#by condition
for(cond in levels(rayleigh_tests$condition))
{
  
print(paste("Rayleigh p<0.05, proportion of", cond, "dances:") )
with(subset(rayleigh_tests,
            subset = condition %in% cond),
     print(mean(p_rayleigh <0.05) *100,digits = 2)
     )
}

```


## Bimodality


## Consistency

### By condition

```{r plot-mean-vector-lengths}
par(mar = c(3, 3, 0, 0))
boxplot(mean_vector~condition,
        data = mean_vectors,
        ylim = c(0,1),
        col = adjustcolor(cond_cols, alpha.f = 0.1),
        border = 'gray50',
        outline = FALSE)
stripchart(mean_vector~condition,
           data = mean_vectors,
           add = TRUE,
           vertical = TRUE,
           method = 'jitter',
           col = adjustcolor(cond_cols, alpha.f = 0.5),
           bg = gray(level = 0,
                     alpha = 0.1),
           pch = 21)
#Identify stimuli
abline(h = c(0,1))
```


```{r summarise-mean-vectors}
mv_agg =  aggregate(mean_vector~condition,
           data = mean_vectors,
           FUN = quantile,
           probs = c(0.5, 0.25, 0.75))
rownames(mv_agg) = with(hd, levels(condition))
mv_agg = within(mv_agg, rm(condition))
mv_agg = t(apply(mv_agg, MARGIN = 1, unlist))
print(mv_agg,
  digits = 3)
```

### By time of day

```{r consistency-time-of-day}
plot(
  mean_vector ~ GetToD(time),
  data = subset(
    mean_vectors,
    condition %in% "vertical"
  ),
  xlim = range(exper_times) + 60*60*c(-1,1),#experiment range +-1h
  # ylim = c(0, 1)+ c(1,0)*0.5,
  xlab = "Time of Day",
  ylab = "Mean vector length",
  main = "Vertical Dance Consistency",
  col = adjustcolor(
    cond_cols$vertical,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = c(0, 1))
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("vertical")
  ),
  {
    lines(
      LogitLoess(
        x = GetToD(time),
        y = mean_vector
      ),

      col = adjustcolor(col = 'darkblue', alpha.f = 0.5),
      lwd = 5
    )
  }
)

```


### By ephemeris error
Could this be a response to ephemeris error? 
Perhaps they are trying to update their ephemeris with uncertainty?

```{r mean-vector-ephemeris-error}
plot(
  mean_vector ~  abs(Mod360.180(
          mu
        )
      ),
  data = subset(
    mean_vectors,
    condition %in% "vertical"
  ),
  xlim = c(0,1)*90,#possible range
  xlab = "Absolute azimuth error",
  ylab = "Mean vector length",
  main = "Vertical Dance Consistency",
  col = adjustcolor(
    cond_cols$vertical,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = c(0, 1))
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("vertical")
  ),
  {
    lines(
      LogitLoess(
        x = abs(Mod360.180(
          mu
        ) ),
        y = mean_vector
      ),

      col = adjustcolor(col = 'darkblue', alpha.f = 0.5),
      lwd = 5
    )
  }
)

```

### By sun elevation


```{r consistency-elevation}
par(mfrow = c(1,2),
    mar = c(4,4,4,0))
plot(
  mean_vector ~ sun_el,
  data = subset(
    mean_vectors,
    condition %in% "vertical"
  ),
  xlim = 90*c(0,1),#possible range
  ylim = c(0, 1)+ c(1,0)*0.01,
  xlab = "Sun elevation",
  ylab = "Mean vector length",
  main = "Vertical Dance Consistency",
  col = adjustcolor(
    cond_cols$vertical,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = c(0, 1))
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("vertical")
  ),
  {
    lines(
      LogitLoess(
        x = sun_el,
        y = mean_vector
      ),

      col = adjustcolor(col = 'darkblue', alpha.f = 0.5),
      lwd = 5
    )
  }
)

#horizontal
plot(
  mean_vector ~ sun_el,
  data = subset(
    mean_vectors,
    condition %in% "horizontal"
  ),
  xlim = 90*c(0,1),#possible range
  ylim = c(0, 1)+ c(1,0)*0.01,
  xlab = "Sun elevation",
  ylab = "",#"Mean vector length",
  main = "Horizontal Dance Consistency",
  col = adjustcolor(
    cond_cols$horizontal,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = c(0, 1))
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("horizontal")
  ),
  {
    lines(
      LogitLoess(
        x = sun_el,
        y = mean_vector
      ),

      col = adjustcolor(col = 'darkblue', alpha.f = 0.5),
      lwd = 5
    )
  }
)

```




## Direction

### By condition

```{r mean-angle-summary}
par(mfrow = c(2,3), 
    mar = c(0,0,0,0),
    oma = c(2,0,0,0))
for(cond in levels(hd$condition))
{
PCfun(angles = subset(mean_vectors, 
                      subset = condition %in% cond)$mu,
      col = cond_cols[[cond]],
      shrink = 2.8,
      title = cond)
}
mtext(text = "mean dance angles relative to feeder",
      side = 1,
      line = 0.2,
      outer = TRUE)

for(tod in c('sun_el < 15', 
             'sun_el < 45 & sun_el > 30',
             'sun_el > 75'
             )
    )
{
  
  # tod_filter = eval(str2lang(tod), envir = mean_vectors)
  for(cond in levels(hd$condition))
{
    cond_tod_mu = subset(mean_vectors, 
                      subset = condition %in% cond &
                        eval(str2lang(tod))
                      )$mu
    if(length(cond_tod_mu)>0)
    {
PCfun(angles = cond_tod_mu,
      col = cond_cols[[cond]],
      shrink = 2.8,
      title = cond)
    }else
    {
       plot(
        NULL,
        axes = FALSE,
        xlim = c(0, 0),
        ylim = c(0, 0),
        xlab = "",
        ylab = "",
        main = ""
      )
      legend(x = "center",
             legend = paste("no", cond, "data found"))
    }
}
mtext(text = paste(tod, "mean dance angles"),
      side = 1,
      line = 0.2,
      outer = TRUE)
  
}

```

### By time of day

Absolute offset by sun elevation.
```{r absolute-error-time}
par(
  mfrow = c(1, 2),
  mar = c(4, 4, 4, 0)
)
plot(
  abs(Mod360.180(mu)) ~ GetToD(time),
  data = subset(
    mean_vectors,
    condition %in% "vertical"
  ),
  xlim = range(exper_times) +
    60 * 60 * c(-1, 1), #experiment range +-1h
  ylim = c(0, 1) * 180,
  xlab = "Time of Day",
  ylab = "Absolute mean angle offset",
  main = "Vertical Dance Accuracy",
  col = adjustcolor(
    cond_cols$vertical,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = 0:1 * c(180), lty = 1) #limits
abline(h = 1:3 * c(45), lty = 2) #gridlines
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("vertical")
  ),
  {
    lines(
      LogitLoess(
        x = GetToD(time),
        y = abs(Mod360.180(mu)),
        absangle = TRUE,
        span = 1.0
      ),

      col = adjustcolor(
        col = 'darkblue',
        alpha.f = 0.5
      ),
      lwd = 5
    )
  }
)

#horizontal
plot(
  abs(Mod360.180(mu)) ~ GetToD(time),
  data = subset(
    mean_vectors,
    condition %in% "horizontal"
  ),
  xlim = range(exper_times) +
    60 * 60 * c(-1, 1), #experiment range +-1h
  ylim = c(0, 1) * 180,
  xlab = "Time of Day",
  ylab = "", #"Mean vector length",
  main = "Horizontal Dance Accuracy",
  col = adjustcolor(
    cond_cols$horizontal,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = 0:1 * c(180), lty = 1) #limits
abline(h = 1:3 * c(45), lty = 2) #gridlines
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("horizontal")
  ),
  {
    lines(
      LogitLoess(
        x = GetToD(time),
        y = abs(Mod360.180(mu)),
        absangle = TRUE,
        span = 1.0
      ),

      col = adjustcolor(
        col = 'darkblue',
        alpha.f = 0.5
      ),
      lwd = 5
    )
  }
)
#antisolar
plot(
  abs(Mod360.180(mu)) ~ GetToD(time),
  data = subset(
    mean_vectors,
    condition %in% "antisolar"
  ),
  xlim = range(exper_times) +
    60 * 60 * c(-1, 1), #experiment range +-1h
  ylim = c(0, 1) * 180,
  xlab = "Time of Day",
  ylab = "Absolute mean angle offset",
  main = "Anti-solar Sky Dance Accuracy",
  col = adjustcolor(
    cond_cols$antisolar,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = 0:1 * c(180), lty = 1) #limits
abline(h = 1:3 * c(45), lty = 2) #gridlines
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("antisolar")
  ),
  {
    lines(
      LogitLoess(
        x = GetToD(time),
        y = abs(Mod360.180(mu)),
        absangle = TRUE,
        span = 1.0
      ),

      col = adjustcolor(
        col = 'darkblue',
        alpha.f = 0.5
      ),
      lwd = 5
    )
  }
)

#solar
plot(
  abs(Mod360.180(mu)) ~ GetToD(time),
  data = subset(
    mean_vectors,
    condition %in% "solar"
  ),
  xlim = range(exper_times) +
    60 * 60 * c(-1, 1), #experiment range +-1h
  ylim = c(0, 1) * 180,
  xlab = "Time of Day",
  ylab = "", #"Mean vector length",
  main = "Solar Sky Dance Accuracy",
  col = adjustcolor(
        cond_cols$solar,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = 0:1 * c(180), lty = 1) #limits
abline(h = 1:3 * c(45), lty = 2) #gridlines
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("solar")
  ),
  {
    lines(
      LogitLoess(
        x = GetToD(time),
        y = abs(Mod360.180(mu)),
        absangle = TRUE,
        span = 1.0
      ),

      col = adjustcolor(
        col = 'darkblue',
        alpha.f = 0.5
      ),
      lwd = 5
    )
  }
)
```

### By sun elevation

```{r mean-angle-elevation-session}
par(
  mfrow = c(1, 3),
  mar = c(0, 0, 0, 0),
  oma = c(2, 0, 0, 0)
)

for (cond in levels(hd$condition)) {
  for (tod in c(
    'sun_el < 15',
    'sun_el < 45 & sun_el > 30',
    'sun_el > 75'
  )) {
    # tod_filter = eval(str2lang(tod), envir = mean_vectors)
    cond_tod_mu = subset(
      mean_vectors,
      subset = condition %in%
        cond &
        eval(str2lang(tod))
    )$mu
    if (length(cond_tod_mu) > 0) {
      PCfun(
        angles = cond_tod_mu,
        col = cond_cols[[cond]],
        shrink = 2.8,
        title = tod
      )
    } else {
      plot(
        NULL,
        axes = FALSE,
        xlim = c(0, 0),
        ylim = c(0, 0),
        xlab = "",
        ylab = "",
        main = ""
      )
      legend(
        x = "center",
        legend = paste(
          "no",
          tod,
          "data found"
        )
      )
    }
  }
    mtext(
      text = paste(
        cond,
        "mean dance angles"
      ),
      side = 1,
      line = 0.2,
      outer = TRUE
    )
}

```

Absolute offset by sun elevation.
```{r absolute-error-sun-elevation}
par(mfrow = c(1,2),
    mar = c(4,4,4,0))
plot(
  abs(Mod360.180(mu)) ~ sun_el,
  data = subset(
    mean_vectors,
    condition %in% "vertical"
  ),
  xlim = 90*c(0,1),#possible range
  ylim = c(0, 1)*180,
  xlab = "Sun elevation",
  ylab = "Absolute mean angle offset",
  main = "Vertical Dance Accuracy",
  col = adjustcolor(
    cond_cols$vertical,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = 0:1 * c(180), lty = 1) #limits
abline(h = 1:3 * c(45), lty = 2) #gridlines
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("vertical")
  ),
  {
    lines(
      LogitLoess(
        x = sun_el,
        y = abs(Mod360.180(mu)),absangle = TRUE
      ),

      col = adjustcolor(col = 'darkblue', alpha.f = 0.5),
      lwd = 5
    )
  }
)

#horizontal
plot(
  abs(Mod360.180(mu)) ~ sun_el,
  data = subset(
    mean_vectors,
    condition %in% "horizontal"
  ),
  xlim = 90*c(0,1),#possible range
  ylim = c(0, 1)*180,
  xlab = "Sun elevation",
  ylab = "",#"Mean vector length",
  main = "Horizontal Dance Accuracy",
  col = adjustcolor(
    cond_cols$horizontal,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = 0:1 * c(180), lty = 1) #limits
abline(h = 1:3 * c(45), lty = 2) #gridlines
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("horizontal")
  ),
  {
    lines(
      LogitLoess(
        x = sun_el,
        y = abs(Mod360.180(mu)),absangle = TRUE
      ),

      col = adjustcolor(col = 'darkblue', alpha.f = 0.5),
      lwd = 5
    )
  }
)
#antisolar
plot(
  abs(Mod360.180(mu)) ~ sun_el,
  data = subset(
    mean_vectors,
    condition %in% "antisolar"
  ),
  xlim = 90*c(0,1),#possible range
  ylim = c(0, 1)*180,
  xlab = "Sun elevation",
  ylab = "Absolute mean angle offset",
  main = "Anti-solar Sky Dance Accuracy",
  col = adjustcolor(
    cond_cols$antisolar,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = 0:1 * c(180), lty = 1) #limits
abline(h = 1:3 * c(45), lty = 2) #gridlines
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("antisolar")
  ),
  {
    lines(
      LogitLoess(
        x = sun_el,
        y = abs(Mod360.180(mu)),absangle = TRUE
      ),

      col = adjustcolor(col = 'darkblue', alpha.f = 0.5),
      lwd = 5
    )
  }
)

#solar
plot(
  abs(Mod360.180(mu)) ~ sun_el,
  data = subset(
    mean_vectors,
    condition %in% "solar"
  ),
  xlim = 90*c(0,1),#possible range
  ylim = c(0, 1)*180,
  xlab = "Sun elevation",
  ylab = "",#"Mean vector length",
  main = "Solar Sky Dance Accuracy",
  col = adjustcolor(
    cond_cols$solar,
    alpha.f = 0.5
  ),
  bg = gray(level = 0, alpha = 0.1),
  pch = 21
)
abline(h = 0:1 * c(180), lty = 1) #limits
abline(h = 1:3 * c(45), lty = 2) #gridlines
#add smooth curve
with(
  subset(
    mean_vectors,
    condition %in% c("solar")
  ),
  {
    lines(
      LogitLoess(
        x = sun_el,
        y = abs(Mod360.180(mu)),absangle = TRUE
      ),

      col = adjustcolor(col = 'darkblue', alpha.f = 0.5),
      lwd = 5
    )
  }
)
```

# Condition contrasts

## Find pairs of dances
Find dances by the same individual within the same experiment for different conditions. 
Ideally these should be close together in time.

```{r sequential-dances-test}
#Test on one individual
id_test = "22b"

mv_subs_test = subset(mean_vectors,
                      ID %in% id_test)

mv_subs_test = within(mv_subs_test,
                      {
                      #find pairs of sequential dances
                      ID_d_pair = c(NA,
                                    sapply(X = unique(ID_d)[-1],
                                         FUN = function(i)
                                           {
                                         paste(ID_d[
                                           which(ID_d %in% i) + c(-1,0)
                                           ],
                                           collapse = " - "
                                           )  
                                         }
                                         )
                                  )
                      #find pairs of sequential conditions
                      cond_pair = c(NA,
                                    sapply(X = unique(ID_d)[-1],
                                         FUN = function(i)
                                           {
                                         paste(condition[
                                           which(ID_d %in% i) + c(-1,0)
                                           ],
                                           collapse = " - "
                                           )  
                                         }
                                         )
                                  )
                      el_diff = c(NA, #1st dance paired with nothing
                                  diff(sun_el)) # sequential differences in sun-elevation
                      day_diff = c(NA,
                                   diff(as.numeric(day)) )
                      is_paired = (abs(el_diff) < 15) & #elevation differs by less than 15°
                                  (day_diff == 0) #on (exactly) the same day
                      }
                      ) 

pairs_test = data.frame(
  do.call(what = rbind,
                     args = with(subset(x = mv_subs_test,
                    subset = is_paired),
                  strsplit(x = unique(ID_d_pair), split = " - ")
                  )
)
)
names(pairs_test) = c("dance1", "dance2")

#find dance conditions
pairs_test = within(pairs_test,
                    {
                      cond1 = sapply(X = dance1,
                                     FUN = function(i)
                                       {
                                       subset(mv_subs_test, ID_d %in% i)$condition
                                     }
                                    )
                      cond2 = sapply(X = dance2,
                                     FUN = function(i)
                                       {
                                       subset(mv_subs_test, ID_d %in% i)$condition
                                     }
                                    )
                    }
                    )
MuPairDiff = function(d1, d2, dta) {
  # print(dta); stop()
        subset(dta, dta$ID_d %in% d2)$mu -
          subset(dta, dta$ID_d %in% d1)$mu
      }
#vertical -> horizontal
vh_test_mu = with(
  subset(
    x = pairs_test,
    cond1 %in%
      "vertical" &
      cond2 %in% "horizontal"
  ),
  {
    mapply(
      FUN = MuPairDiff,
      d1 = dance1,
      d2 = dance2,
      dta = list(mean_vectors)
    )
  }
)
#horizontal -> vertical
hv_test_mu = with(
  subset(
    x = pairs_test,
    cond1 %in%
      "horizontal" &
      cond2 %in% "vertical"
  ),
  {
    mapply(
      FUN = MuPairDiff,
      d1 = dance1,
      d2 = dance2,
      dta = list(mean_vectors)
    )
  }
)
#vertical -> vertical
vv_test_mu = with(
  subset(
    x = pairs_test,
    cond1 %in%
      "vertical" &
      cond2 %in% "vertical"
  ),
  {
    mapply(
      FUN = MuPairDiff,
      d1 = dance1,
      d2 = dance2,
      dta = list(mean_vectors)
    )
  }
)
#horizontal -> antisolar
ha_test_mu = with(
  subset(
    x = pairs_test,
    cond1 %in%
      "horizontal" &
      cond2 %in% "antisolar"
  ),
  {
    mapply(
      FUN = MuPairDiff,
      d1 = dance1,
      d2 = dance2,
      dta = list(mean_vectors)
    )
  }
)
#antisolar -> horizontal
ah_test_mu = with(
  subset(
    x = pairs_test,
    cond1 %in%
      "antisolar" &
      cond2 %in% "horizontal"
  ),
  {
    mapply(
      FUN = MuPairDiff,
      d1 = dance1,
      d2 = dance2,
      dta = list(mean_vectors)
    )
  }
)
#antisolar -> solar
as_test_mu = with(
  subset(
    x = pairs_test,
    cond1 %in%
      "antisolar" &
      cond2 %in% "solar"
  ),
  {
    mapply(
      FUN = MuPairDiff,
      d1 = dance1,
      d2 = dance2,
      dta = list(mean_vectors)
    )
  }
)
#solar -> antisolar
sa_test_mu = with(
  subset(
    x = pairs_test,
    cond1 %in%
      "solar" &
      cond2 %in% "antisolar"
  ),
  {
    mapply(
      FUN = MuPairDiff,
      d1 = dance1,
      d2 = dance2,
      dta = list(mean_vectors)
    )
  }
)

  par(
    mfrow = c(2, 2),
    mar = c(0, 0, 0, 0),
    oma = c(1, 0, 0, 0)
  )
PCfun(vv_test_mu,
      col = cond_cols$vertical,
      title = "Vertical - Vertical")
# PCfun(c(vh_test_mu, -hv_test_mu),
#       col = cond_cols$horizontal,
#       title = "Horizontal - Vertical")
PCfun(c(vh_test_mu),
      col = cond_cols$horizontal,
      title = "Horizontal - Vertical")
# PCfun(c(ha_test_mu,-ah_test_mu),
#       col = cond_cols$antisolar,
#       title = "Antisolar - Horizontal")
PCfun(c(ha_test_mu),
      col = cond_cols$antisolar,
      title = "Antisolar - Horizontal")
# PCfun(c(as_test_mu,-sa_test_mu),
#       col = cond_cols$solar,
#       title = "Solar - Antisolar")
PCfun(c(as_test_mu),
      col = cond_cols$solar,
      title = "Solar - Antisolar")
mtext(text = "Sequential mean direction changes, B22",
           side = 1,
      line = -0.25,
      outer = TRUE)

```


Since mean vectors should be organised by calendar time by default, 
sorting them by individual will allow us to determine the sequential differences.

These can then be subset according to individual, and reorganised into true sequential pairs.

```{r all-sequential-dances-test}
#Sort mean vectors by individual

mean_vectors = mean_vectors[order(mean_vectors$ID), ]#sort by individual

#check the date ordering
#are calendar dates in increasing order
#check for an individual with lots of dances
with(
  subset(mean_vectors, ID %in% "22B"),
  {
    #calculate calendar dates and take sequential differences
    tdiff = diff(as.POSIXct(
      x = paste(day, time),
      tz = "Africa/Johannesburg",
      format = "%Y%m%d %H:%M:%OS"
    ))
    if (all(!(tdiff < 0))) { #N.B. some videos were close and are labelled with the same time
      print(
        "For B22, all dance times are in increasing order"
      )
    } else {
      print(paste(
        paste(
          ID_d[
            which(tdiff < 0) +
              c(0, 1)
          ],
          collapse = " & "
        ),
        "are not in order!"
      ))
    }
  }
)

#Set up data.frame with 1st and 2nd dance details
ID_d1 = with(mean_vectors, ID_d[1:(length(u_id_d)-1)])#1st dances
ID_d2 = with(mean_vectors, ID_d[2:length(u_id_d)])#2nd dances
#align information about the pairs
mv_pairs = data.frame( cbind( 
                  subset(mean_vectors, ID_d %in% ID_d1),
                  subset(mean_vectors, ID_d %in% ID_d2)
                  )
                ) 
#names(mv_pairs) #the names for the 2nd dance columns now end in ".1"

#select only dance pairs by the same individual
mv_pairs = subset(mv_pairs, ID == ID.1)

#add columns for differences
mv_pairs = within(mv_pairs,
                  {
                    time_diff = as.POSIXct(
                                  x = paste(day.1, time.1),
                                  tz = "Africa/Johannesburg",
                                  format = "%Y%m%d %H:%M:%OS"
                                ) - as.POSIXct(
                                  x = paste(day, time),
                                  tz = "Africa/Johannesburg",
                                  format = "%Y%m%d %H:%M:%OS"
                                )#difference in time in seconds
                    ele_diff = sun_el.1 - sun_el
                    az_diff = Mod360.180(sun_az.1 - sun_az)
                    mv_diff = mean_vector.1 - mean_vector
                    kp_diff = kappa.1 - kappa
                    mu_diff = Mod360.180(mu.1 - mu)
                    
                  }
                  )
#test again the all dances are sequential
with(
  mv_pairs,
  {
    if (all(!(time_diff < 0))) {
      print(
        "All dance pair times are in increasing order."
      )
    } else {
      message(paste(
        paste(
          ID_d[
            which(time_diff < 0)
          ],
          ID_d.1[
            which(time_diff < 0)
          ],
          "\n"
        ),
        "are not in order!\n"
      ))
    }
  }
)


mv_subs = subset(mv_pairs,
                      (abs(ele_diff) < 15) & #elevation differs by less than 15°
                                  (time_diff < 30*60) #less than 30 minutes
                      ) 
```

## Direction

```{r, dance-pairs-controls}
par(
  mfrow = c(2, 3),
  mar = c(0, 0, 0, 0),
  oma = c(1, 0, 0, 0)
)
#all controls
PCfun(PairedDiff(dta = mv_subs,
  cond_1 = "vertical",
  cond_2 = "vertical"),
      shrink = 2.0,
      col = cond_cols$vertical,
      title = "Vertical - Vertical")

PCfun(PairedDiff(dta = mv_subs,
  cond_1 = "horizontal",
  cond_2 = "horizontal"),
      shrink = 2.0,
      col = cond_cols$horizontal,
      title = "Horizontal - Horizontal")
PCfun(PairedDiff(dta = mv_subs,
  cond_1 = "tilted",
  cond_2 = "tilted"),
      shrink = 2.0,
      col = cond_cols$tilted,
      title = "Tilted - Tilted")
PCfun(PairedDiff(dta = mv_subs,
  cond_1 = "antisolar",
  cond_2 = "antisolar"),
      shrink = 2.0,
      col = cond_cols$antisolar,
      title = "Antisolar - Antisolar")
PCfun(PairedDiff(dta = mv_subs,
  cond_1 = "solar",
  cond_2 = "solar"),
      shrink = 2.0,
      col = cond_cols$solar,
      title = "Solar - Solar")
PCfun(PairedDiff(dta = mv_subs,
  cond_1 = "zenith",
  cond_2 = "zenith"),
      shrink = 2.0,
      col = cond_cols$zenith,
      title = "Zenithal - Zenithal")
mtext(text = "Sequential mean direction changes, controls, all bees",
           side = 1,
      line = -0.25,
      outer = TRUE)
```


```{r, condition-pair-comparisons}

par(
  mfrow = c(2, 3),
  mar = c(0, 0, 0, 0),
  oma = c(1, 0, 0, 0)
)

PCfun(angles = PairedDiff(dta = 
  mv_subs,
  cond_1 = "vertical",
  cond_2 = "horizontal"
),
      shrink = 2.0,
      col = cond_cols$horizontal,
      title = "Horizontal - Vertical")

PCfun(angles =  PairedDiff(dta = 
  mv_subs,
  cond_1 = "horizontal",
  cond_2 = "antisolar"
),
      shrink = 2.0,
      col = cond_cols$antisolar,
      title = "Antisolar - Horizontal")

PCfun(angles =  PairedDiff(dta = 
  mv_subs,
  cond_1 = "horizontal",
  cond_2 = "solar"
),
      shrink = 2.0,
      col = cond_cols$solar,
      title = "Solar - Horizontal")

PCfun(angles = PairedDiff(dta = 
  mv_subs,
  cond_1 = "antisolar",
  cond_2 = "solar"
),
      shrink = 2.0,
      col = cond_cols$solar,
      title = "Solar - Antisolar")

PCfun(PairedDiff(dta = 
  mv_subs,
  cond_1 = "horizontal",
  cond_2 = "zenithal"
),
      shrink = 2.0,
      col = cond_cols$zenith,
      title = "Zenithal - Horizontal")

mtext(text = "Sequential mean direction changes, all bees",
           side = 1,
      line = -0.25,
      outer = TRUE)
```


### By sun elevation
```{r, dance-pairs-low}
mv_low = subset(mv_subs,
                sun_el <15)

par(
  mfrow = c(2, 3),
  mar = c(0, 0, 0, 0),
  oma = c(1, 0, 0, 0)
)
#all controls
PCfun(PairedDiff(dta = mv_low,
  cond_1 = "vertical",
  cond_2 = "vertical"),
      col = cond_cols$vertical,
      title = "Vertical - Vertical")

PCfun(PairedDiff(dta = mv_low,
  cond_1 = "horizontal",
  cond_2 = "horizontal"),
      col = cond_cols$horizontal,
      title = "Horizontal - Horizontal")
PCfun(PairedDiff(dta = mv_low,
  cond_1 = "tilted",
  cond_2 = "tilted"),
      col = cond_cols$tilted,
      title = "Tilted - Tilted")
PCfun(PairedDiff(dta = mv_low,
  cond_1 = "antisolar",
  cond_2 = "antisolar"),
      col = cond_cols$antisolar,
      title = "Antisolar - Antisolar")
PCfun(PairedDiff(dta = mv_low,
  cond_1 = "solar",
  cond_2 = "solar"),
      col = cond_cols$solar,
      title = "Solar - Solar")
PCfun(PairedDiff(dta = mv_low,
  cond_1 = "zenith",
  cond_2 = "zenith"),
      col = cond_cols$zenith,
      title = "Zenithal - Zenithal")
mtext(text = "Low elevation, controls, all bees",
           side = 1,
      line = -0.25,
      outer = TRUE)

PCfun(angles = PairedDiff(dta = 
  mv_low,
  cond_1 = "vertical",
  cond_2 = "horizontal"
),
      col = cond_cols$horizontal,
      title = "Horizontal - Vertical")

PCfun(angles =  PairedDiff(dta = 
  mv_low,
  cond_1 = "horizontal",
  cond_2 = "antisolar"
),
      col = cond_cols$antisolar,
      title = "Antisolar - Horizontal")

PCfun(angles =  PairedDiff(dta = 
  mv_low,
  cond_1 = "horizontal",
  cond_2 = "solar"
),
      col = cond_cols$solar,
      title = "Solar - Horizontal")

PCfun(angles = PairedDiff(dta = 
  mv_low,
  cond_1 = "antisolar",
  cond_2 = "solar"
),
      col = cond_cols$solar,
      title = "Solar - Antisolar")

PCfun(PairedDiff(dta = 
  mv_low,
  cond_1 = "horizontal",
  cond_2 = "zenithal"
),
      col = cond_cols$zenith,
      title = "Zenithal - Horizontal")

mtext(text = "Low elevation changes, all bees",
           side = 1,
      line = -0.25,
      outer = TRUE)

```


```{r, dance-pairs-medium}
mv_med = subset(mv_subs,
                sun_el >30 & sun_el <45)

par(
  mfrow = c(2, 3),
  mar = c(0, 0, 0, 0),
  oma = c(1, 0, 0, 0)
)
#all controls
PCfun(PairedDiff(dta = mv_med,
  cond_1 = "vertical",
  cond_2 = "vertical"),
      col = cond_cols$vertical,
      title = "Vertical - Vertical")

PCfun(PairedDiff(dta = mv_med,
  cond_1 = "horizontal",
  cond_2 = "horizontal"),
      col = cond_cols$horizontal,
      title = "Horizontal - Horizontal")
PCfun(PairedDiff(dta = mv_med,
  cond_1 = "tilted",
  cond_2 = "tilted"),
      col = cond_cols$tilted,
      title = "Tilted - Tilted")
PCfun(PairedDiff(dta = mv_med,
  cond_1 = "antisolar",
  cond_2 = "antisolar"),
      col = cond_cols$antisolar,
      title = "Antisolar - Antisolar")
PCfun(PairedDiff(dta = mv_med,
  cond_1 = "solar",
  cond_2 = "solar"),
      col = cond_cols$solar,
      title = "Solar - Solar")
PCfun(PairedDiff(dta = mv_med,
  cond_1 = "zenith",
  cond_2 = "zenith"),
      col = cond_cols$zenith,
      title = "Zenithal - Zenithal")
mtext(text = "Medium elevation, controls, all bees",
           side = 1,
      line = -0.25,
      outer = TRUE)

PCfun(angles = PairedDiff(dta = 
  mv_med,
  cond_1 = "vertical",
  cond_2 = "horizontal"
),
      col = cond_cols$horizontal,
      title = "Horizontal - Vertical")

PCfun(angles =  PairedDiff(dta = 
  mv_med,
  cond_1 = "horizontal",
  cond_2 = "antisolar"
),
      col = cond_cols$antisolar,
      title = "Antisolar - Horizontal")

PCfun(angles =  PairedDiff(dta = 
  mv_med,
  cond_1 = "horizontal",
  cond_2 = "solar"
),
      col = cond_cols$solar,
      title = "Solar - Horizontal")

PCfun(angles = PairedDiff(dta = 
  mv_med,
  cond_1 = "antisolar",
  cond_2 = "solar"
),
      col = cond_cols$solar,
      title = "Solar - Antisolar")

PCfun(PairedDiff(dta = 
  mv_med,
  cond_1 = "horizontal",
  cond_2 = "zenithal"
),
      col = cond_cols$zenith,
      title = "Zenithal - Horizontal")

mtext(text = "Medium elevation changes, all bees",
           side = 1,
      line = -0.25,
      outer = TRUE)

```


```{r, dance-pairs-high}
mv_hig = subset(mv_subs,
                sun_el >75)

par(
  mfrow = c(2, 3),
  mar = c(0, 0, 0, 0),
  oma = c(1, 0, 0, 0)
)
#all controls
PCfun(PairedDiff(dta = mv_hig,
  cond_1 = "vertical",
  cond_2 = "vertical"),
      col = cond_cols$vertical,
      title = "Vertical - Vertical")

PCfun(PairedDiff(dta = mv_hig,
  cond_1 = "horizontal",
  cond_2 = "horizontal"),
      col = cond_cols$horizontal,
      title = "Horizontal - Horizontal")
PCfun(PairedDiff(dta = mv_hig,
  cond_1 = "tilted",
  cond_2 = "tilted"),
      col = cond_cols$tilted,
      title = "Tilted - Tilted")
PCfun(PairedDiff(dta = mv_hig,
  cond_1 = "antisolar",
  cond_2 = "antisolar"),
      col = cond_cols$antisolar,
      title = "Antisolar - Antisolar")
PCfun(PairedDiff(dta = mv_hig,
  cond_1 = "solar",
  cond_2 = "solar"),
      col = cond_cols$solar,
      title = "Solar - Solar")
PCfun(PairedDiff(dta = mv_hig,
  cond_1 = "zenith",
  cond_2 = "zenith"),
      col = cond_cols$zenith,
      title = "Zenithal - Zenithal")
mtext(text = "High elevation, controls, all bees",
           side = 1,
      line = -0.25,
      outer = TRUE)

PCfun(angles = PairedDiff(dta = 
  mv_hig,
  cond_1 = "vertical",
  cond_2 = "horizontal"
),
      col = cond_cols$horizontal,
      title = "Horizontal - Vertical")

PCfun(angles =  PairedDiff(dta = 
  mv_hig,
  cond_1 = "horizontal",
  cond_2 = "antisolar"
),
      col = cond_cols$antisolar,
      title = "Antisolar - Horizontal")

PCfun(angles =  PairedDiff(dta = 
  mv_hig,
  cond_1 = "horizontal",
  cond_2 = "solar"
),
      col = cond_cols$solar,
      title = "Solar - Horizontal")

PCfun(angles = PairedDiff(dta = 
  mv_hig,
  cond_1 = "antisolar",
  cond_2 = "solar"
),
      col = cond_cols$solar,
      title = "Solar - Antisolar")

PCfun(PairedDiff(dta = 
  mv_hig,
  cond_1 = "horizontal",
  cond_2 = "zenithal"
),
      col = cond_cols$zenith,
      title = "Zenithal - Horizontal")

mtext(text = "High elevation changes, all bees",
           side = 1,
      line = -0.25,
      outer = TRUE)

```