---
title: "Exploratory Plots"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    fig_caption: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
  # bookdown::pdf_document2:
  #   highlight: rstudio
---

# References


# TODO

- Load data +
- Reformat +
- Select dances of interest
- Plot dances of interest
- Search for more examples
- Plot additional examples

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load functions and packages

Load the packages and other functions that will be used during reorganisation.

```{r load-functions}
source('HS_functions.R')
```


# Load data


## Load the raw data

Load the raw data exported from Excel to `csv`.

```{r load-the-data}
path_file = "0Data/hs_dance_reorg.csv"
if(!file.exists(path_file))
{print(path_file, '\nnot found')}
hd = read.table(file = path_file, 
                header = T, 
                sep  = ',')
```


## Inspect data

Check the names of variables in this dataset.

```{r check-names}
names(hd)
```

<details>

<summary>Click to expand raw data</summary>

```{r inspect-data}
print(hd)
```


# Format data

Ensure that each variable has the correct data format for subsequent analysis.

```{r Data formatting}
hd = within(hd,
            {
              ID = as.factor(ID) # bee identifier as a factor
              ID_d = as.factor(ID_d) # beedance identifier as a factor
              day = as.factor(day) # date as a factor
              sast_time = as.POSIXct(sast_time,
                                     tz = "Africa/Johannesburg", 
                                      format = "%Y-%m-%d %H:%M:%OS")#South African Standard time
            }
)
u_id_d = with(hd, unique(ID_d)) # unique beedances
length(u_id_d)#68 beedances
```

# Plot Examples


### Plot an example dance

For example we can look at the dances for Blue-22 recorded on 2025-01-28 at 
07:57:05, 08:10:00, 08:29:00, 08:34:00 & 10:44:09—each of which is the first for 
one of the stimulus conditions.
Sun azimuth (possibly incorrect) is shown as an orange line.

```{r example-subset}
subs_20250128 = subset(hd, day %in% "2025-01-28")
#find first dance for each condition on 2025.01.28
example_lst = list()
for(stim in unique(hd$stimulus))
{
example_lst[[stim]] = with(subset(subs_20250128, stimulus %in% stim),
                           unique(ID_d)[1])
}
#inspect

example_subs = subset(x = hd,
                      ID_d %in% paste(unlist(example_lst))
                      )

par(mfrow = c(2,2), mar = c(0,0,0,0))
with(subset(example_subs,
            subset = stimulus %in% 'vertical'),
     {
  PCfun(angles = bearing,
        col = 'darkgreen',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
  lines(
    x = c(0, sin(rad(feeder_az))),
    y = c(0, cos(rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)

with(subset(example_subs,
            subset = stimulus %in% 'horizontal'),
     {
PCfun(angles = bearing,
      col = 'darkred',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
         lines(
    x = c(0, sin(rad(feeder_az))),
    y = c(0, cos(rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
with(subset(example_subs,
            subset = stimulus %in% 'antisolar'),
     {
PCfun(angles = bearing,
      col = 'cyan4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
         lines(
    x = c(0, sin(rad(feeder_az))),
    y = c(0, cos(rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
with(subset(example_subs,
            subset = stimulus %in% 'solar'),
     {
PCfun(angles = bearing,
      col = 'blue4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
         lines(
    x = c(0, sin(rad(feeder_az+180))),
    y = c(0, cos(rad(feeder_az+180))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
```


These should look like the dances above but rotated so that the feeder direction 
falls at 0°. I suspect there is a problem with angles in incompatible formats 
in the data organisation script.

```{r feeder-angle-examples}


par(mfrow = c(2,2), mar = c(0,0,0,0))
with(subset(example_subs,
            subset = stimulus %in% 'vertical'),
     {
  PCfun(angles = deg(feeder_angle),
        col = 'darkgreen',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
     }
)

with(subset(example_subs,
            subset = stimulus %in% 'horizontal'),
     {
PCfun(angles = deg(feeder_angle),
      col = 'darkred',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
     }
)
with(subset(example_subs,
            subset = stimulus %in% 'antisolar'),
     {
PCfun(angles = deg(feeder_angle),
      col = 'cyan4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
     }
)
with(subset(example_subs,
            subset = stimulus %in% 'solar'),
     {
PCfun(angles = deg(feeder_angle),
      col = 'blue4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
     }
)
```


## Low elevation


```{r low-elevation-examples}
#find first dance for each condition on 2025.01.28
lowel_lst = list()
for(stim in unique(hd$stimulus))
{
lowel_lst[[stim]] = with(subset(hd, stimulus %in% stim &
                                  sun_el < 15),
                           unique(ID_d))
}
#inspect
sapply(lowel_lst, length)# no low elevation for dome 


lowel_subs = subset(x = hd,
                      ID_d %in% paste(unlist(lowel_lst))
                      )

par(mfrow = c(2,4), mar = c(0,0,0,0))
for(dnc in lowel_lst[['vertical']])
{
with(subset(lowel_subs,
            subset = stimulus %in% 'vertical' &
              ID_d %in% dnc
              ),
     {
  PCfun(angles = deg(feeder_angle),
        col = 'darkgreen',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
  # lines(
  #   x = c(0, sin(rad(feeder_az))),
  #   y = c(0, cos(rad(feeder_az))),
  #   col = adjustcolor('orange', alpha.f = 0.5),
  #   lwd = 2
  # )
     }
)
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in lowel_lst[['horizontal']]) {
  with(
    subset(lowel_subs, subset = stimulus %in% 'horizontal' & ID_d %in% dnc),
    {
      PCfun(
        angles = deg(feeder_angle),
        col = 'darkred',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
      #        lines(
      #   x = c(0, sin(rad(feeder_az))),
      #   y = c(0, cos(rad(feeder_az))),
      #   col = adjustcolor('orange', alpha.f = 0.5),
      #   lwd = 2
      # )
    }
  )
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in lowel_lst[['antisolar']]) {
  with(subset(lowel_subs, subset = stimulus %in% 'antisolar' & ID_d %in% dnc), {
    PCfun(
      angles = deg(feeder_angle),
      col = 'cyan4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
    #        lines(
    #   x = c(0, sin(rad(feeder_az))),
    #   y = c(0, cos(rad(feeder_az))),
    #   col = adjustcolor('orange', alpha.f = 0.5),
    #   lwd = 2
    # )
  })
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in lowel_lst[['solar']]) {
  with(subset(lowel_subs, subset = stimulus %in% 'solar' & ID_d %in% dnc), {
    PCfun(
      angles = deg(feeder_angle),
      col = 'blue4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
    #        lines(
    #   x = c(0, sin(rad(feeder_az+180))),
    #   y = c(0, cos(rad(feeder_az+180))),
    #   col = adjustcolor('orange', alpha.f = 0.5),
    #   lwd = 2
    # )
  })
}
```



## Medium elevation


```{r low-elevation-examples}
#dances for each condition
medel_lst = list()
for(stim in unique(hd$stimulus))
{
medel_lst[[stim]] = with(subset(hd, stimulus %in% stim &
                                  sun_el > 30 &
                                  sun_el < 45 
                                  ),
                           unique(ID_d))
}
#inspect
sapply(medel_lst, length)# no low elevation for dome 


medel_subs = subset(x = hd,
                      ID_d %in% paste(unlist(medel_lst))
                      )

par(mfrow = c(2,4), mar = c(0,0,0,0))
for(dnc in medel_lst[['vertical']])
{
with(subset(medel_subs,
            subset = stimulus %in% 'vertical' &
              ID_d %in% dnc
              ),
     {
  PCfun(angles = deg(feeder_angle),
        col = 'darkgreen',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
  # lines(
  #   x = c(0, sin(rad(feeder_az))),
  #   y = c(0, cos(rad(feeder_az))),
  #   col = adjustcolor('orange', alpha.f = 0.5),
  #   lwd = 2
  # )
     }
)
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in medel_lst[['horizontal']]) {
  with(
    subset(medel_subs, subset = stimulus %in% 'horizontal' & ID_d %in% dnc),
    {
      PCfun(
        angles = deg(feeder_angle),
        col = 'darkred',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
      #        lines(
      #   x = c(0, sin(rad(feeder_az))),
      #   y = c(0, cos(rad(feeder_az))),
      #   col = adjustcolor('orange', alpha.f = 0.5),
      #   lwd = 2
      # )
    }
  )
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in medel_lst[['antisolar']]) {
  with(subset(medel_subs, subset = stimulus %in% 'antisolar' & ID_d %in% dnc), {
    PCfun(
      angles = deg(feeder_angle),
      col = 'cyan4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
    #        lines(
    #   x = c(0, sin(rad(feeder_az))),
    #   y = c(0, cos(rad(feeder_az))),
    #   col = adjustcolor('orange', alpha.f = 0.5),
    #   lwd = 2
    # )
  })
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in medel_lst[['solar']]) {
  with(subset(medel_subs, subset = stimulus %in% 'solar' & ID_d %in% dnc), {
    PCfun(
      angles = deg(feeder_angle),
      col = 'blue4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
    #        lines(
    #   x = c(0, sin(rad(feeder_az+180))),
    #   y = c(0, cos(rad(feeder_az+180))),
    #   col = adjustcolor('orange', alpha.f = 0.5),
    #   lwd = 2
    # )
  })
}
```




## High elevation


```{r low-elevation-examples}
#dances for each condition
higel_lst = list()
for(stim in unique(hd$stimulus))
{
higel_lst[[stim]] = with(subset(hd, stimulus %in% stim &
                                  sun_el > 75 
                                  ),
                           unique(ID_d))
}
#inspect
sapply(higel_lst, length)# no low elevation for dome 


higel_subs = subset(x = hd,
                      ID_d %in% paste(unlist(higel_lst))
                      )

par(mfrow = c(2,4), mar = c(0,0,0,0))
for(dnc in higel_lst[['vertical']])
{
with(subset(higel_subs,
            subset = stimulus %in% 'vertical' &
              ID_d %in% dnc
              ),
     {
  PCfun(angles = deg(feeder_angle),
        col = 'darkgreen',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
  # lines(
  #   x = c(0, sin(rad(feeder_az))),
  #   y = c(0, cos(rad(feeder_az))),
  #   col = adjustcolor('orange', alpha.f = 0.5),
  #   lwd = 2
  # )
     }
)
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in higel_lst[['horizontal']]) {
  with(
    subset(higel_subs, subset = stimulus %in% 'horizontal' & ID_d %in% dnc),
    {
      PCfun(
        angles = deg(feeder_angle),
        col = 'darkred',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
      #        lines(
      #   x = c(0, sin(rad(feeder_az))),
      #   y = c(0, cos(rad(feeder_az))),
      #   col = adjustcolor('orange', alpha.f = 0.5),
      #   lwd = 2
      # )
    }
  )
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in higel_lst[['antisolar']]) {
  with(subset(higel_subs, subset = stimulus %in% 'antisolar' & ID_d %in% dnc), {
    PCfun(
      angles = deg(feeder_angle),
      col = 'cyan4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
    #        lines(
    #   x = c(0, sin(rad(feeder_az))),
    #   y = c(0, cos(rad(feeder_az))),
    #   col = adjustcolor('orange', alpha.f = 0.5),
    #   lwd = 2
    # )
  })
}

par(mfrow = c(2,4), mar = c(0,0,0,0))
for (dnc in higel_lst[['solar']]) {
  with(subset(higel_subs, subset = stimulus %in% 'solar' & ID_d %in% dnc), {
    PCfun(
      angles = deg(feeder_angle),
      col = 'blue4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus), "\n", unique(ID)))
    #        lines(
    #   x = c(0, sin(rad(feeder_az+180))),
    #   y = c(0, cos(rad(feeder_az+180))),
    #   col = adjustcolor('orange', alpha.f = 0.5),
    #   lwd = 2
    # )
  })
}
```