---
title: "Organise Data"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    fig_caption: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
  # bookdown::pdf_document2:
  #   highlight: rstudio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# References

> Natalie Cooper & Pen-Yuan Hsing, 2017 [A Guide to Reprhducible Chde in Ecology and Evolution](https://britishecologicalsociety.org/wp-content/uploads/2017/12/guide-to-reprhducible-chde.pdf) British Ecological Society

# ThdO

- Load data +
- Reorganise +
- Add solar azimuth +
- Export +
- Location of cel_bhdy +

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load functions

Load the packages that will be used during reorganisation.

```{r load-packages}
suppressMessages(#these are disturbing users unnecessarily
  {
    require(circular)#package for handling circular data
    require(oce)#package for calculating sun azimuth
  }
)
```

Load other functions relevant to this project.

```{r load-functions}
source('HS_functions.R')
```

# Load data

## Load the raw data

Load the raw data exported from Excel to `csv`.

```{r load-the-data}
path_file = "0Data/B22_ waggle_run_directions.csv"
if(!file.exists(path_file))
{print(path_file, '\nnot found')}
hd = read.table(file = path_file, 
                header = T, 
                sep  = ',')
```

## Inspect data


<details>

<summary>Click to expand raw data</summary>

```{r inspect-data}
print(hd)
```

# Reorganise

## Rename variables

Rename some variables.


```{r rename-variables}
hd = within(hd, {
  ID = paste0(dancer_id) # individual identifier
  ID_d = paste0(dance_id) # dance identifier
  date = source_batch # this is already in POSIX format
  #clean up information that has been doubled
  rm(list = c('dancer_id', 'dance_id', 'dance_key', 'source_batch', "date",
              "waggle_directions",#split into direction_u  and direction_v
              "start_xy"#split into "start_x" and "start_y" 
              ))
})
```


```{r summarise-factors}
nm_fact = c( "ID",         
 "ID_d"          ,
 # "dance_key"         ,
 # [4,] "waggle_run"        
 "stimulus"         , 
 # "notes"            , 
 # [7,] "waggle_directions" 
#  [8,] "start_x"           
#  [9,] "start_y"           
# [10,] "start_xy"          
# [11,] "csv_file"          
# [12,] "time"              
# [13,] "date"              
# [14,] "direction_u"       
# [15,] "direction_v"       
# [16,] "video_angle_deg"   
# [17,] "start_frame"       
 "video_name"        
# [19,] "annotation_session"
# [20,] "source_batch"
            )
sapply(X = hd[, nm_fact],
       unique
     )
```

## Reformat variables

Reformat variables to factor and numeric. Everything else remains character format by default.


```{r reformat-variables}
nm_num = c(
 #  "dancer_id",         
 # "dance_id"          ,
 # "dance_key"         ,
 "waggle_run",        
 # "stimulus"         , 
 # "notes"            , 
 # [7,] "waggle_directions" 
"start_x"    ,       
"start_y"     ,      
# "start_xy"          
# [11,] "csv_file"          
# [12,] "time"              
# [13,] "date"              
"direction_u",
"direction_v",
"video_angle_deg",
"start_frame",
 # "video_name"        
"annotation_session"
# [20,] "source_batch"
)
#format factors
for(nm in nm_fact)
{
hd[, nm] = as.factor(hd[, nm])
}

#format numeric
for(nm in nm_num)
{
hd[, nm] = as.numeric(hd[, nm])
}

summary(hd)
```

There appears to have been an autofill error in stimulus labelling. 
Rename all unexpected stimuli to `deg_10`, 10° above horizontal.
```{r reformat-stimulus}
hd = within(hd,
            {
              stimulus = sapply(as.character(stimulus),
                                switch,
                                vertical = "vertical",
                                horizontal = "horizontal",
                                antisolar = "antisolar",
                                solar = "solar",
                                deg_10 = "deg_10",
                                "deg_10")#otherwise always deg_10
            }
            )
with(hd, unique(stimulus))
```



## Format angles

```{r format-angles}
#assign angles to numeric

#check angle format
summary(
  with(hd, {
    atan2(y = direction_v, x = direction_u) - rad(video_angle_deg) + pi / 2
  })
)
#video_angle_deg is the dance angle rel to vertical, except for the occasional rounding error
hd = within(hd, 
            {
              signed_angle = Mod360.180(video_angle_deg)  # bearing between -180 and 180
              angle = circular(rad(signed_angle),# bearing between -pi and pi
                               rotation = 'clock') # circular format suppresses later warnings
            }
            )
with(hd, summary(angle))
```


