---
title: "Organise Data"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    fig_caption: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
  # bookdown::pdf_document2:
  #   highlight: rstudio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# References

> Natalie Cooper & Pen-Yuan Hsing, 2017 [A Guide to Reprhducible Chde in Ecology and Evolution](https://britishecologicalsociety.org/wp-content/uploads/2017/12/guide-to-reprhducible-chde.pdf) British Ecological Society

# ThdO

- Load data +
- Reorganise +
- Add solar azimuth +
- Export +
- Location of cel_bhdy +

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load functions

Load the packages that will be used during reorganisation.

```{r load-packages}
suppressMessages(#these are disturbing users unnecessarily
  {
    require(circular)#package for handling circular data
    require(oce)#package for calculating sun azimuth
  }
)
```

Load other functions relevant to this project.

```{r load-functions}
source('HS_functions.R')
```

# Load data

## Load the raw data

Load the raw data exported from Excel to `csv`.

```{r load-the-data}
path_file = "0Data/B22_ waggle_run_directions.csv"
if(!file.exists(path_file))
{print(path_file, '\nnot found')}
hd = read.table(file = path_file, 
                header = T, 
                sep  = ',')
```

## Inspect data

This data is in wide format, and combines multiple condition factors into single columns.
We will want to convert to long format, identifying the trial number for each exit, and to split condition labels into their constituent parts.

```{r summarise-factors}
nm_fact = c( "dancer_id",         
 "dance_id"          ,
 "dance_key"         ,
 # [4,] "waggle_run"        
 "stimulus"         , 
 # "notes"            , 
 # [7,] "waggle_directions" 
#  [8,] "start_x"           
#  [9,] "start_y"           
# [10,] "start_xy"          
# [11,] "csv_file"          
# [12,] "time"              
# [13,] "date"              
# [14,] "direction_u"       
# [15,] "direction_v"       
# [16,] "video_angle_deg"   
# [17,] "start_frame"       
 "video_name"        
# [19,] "annotation_session"
# [20,] "source_batch"
            )
sapply(X = hd[, nm_fact],
       unique
     )
```


<details>

<summary>Click to expand raw data</summary>

```{r inspect-data}
print(hd)
```

# Reorganise

## Rename variables

Rename some variables.


### Rename variables

```{r rename-variables}
hd = within(od, {
  ID = paste0(Beetle_Id) # individual identifier
  date = gsub(
    pattern = '[.:]',
    #most dates separated with ".", but one is separated with ":"
    replacement = '-',
    x = Day
  ) #converts to a date in POSIX format
  Species = ifelse(Species %in% 'K.lamercki', yes = 'K.lamarcki', no = Species)
  Latitude = as.numeric(gsub(
    pattern = ",.*",
    #replace after the comma
    replacement = "",
    x = Coordinates
  ))
  Longitude = as.numeric(gsub(
    pattern = ".*,\\s*",
    #replace before and any spaces (\s*) immediately after
    replacement = "",
    x = Coordinates
  ))
  #relace sex "" with NA
  Sex = ifelse(test = Sex %in% c("m", "f"),
                         yes = Sex,
                         no = NA)
  #clean up information that has been doubled
  rm(list = c('Beetle_Id', 'Day', 'Year', "Coordinates"))
})
```


