---
title: "Organise Data"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    fig_caption: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
  # bookdown::pdf_document2:
  #   highlight: rstudio
---

# References

> Natalie Cooper & Pen-Yuan Hsing, 2017 [A Guide to Reproducible Code in Ecology and Evolution](https://britishecologicalsociety.org/wp-content/uploads/2017/12/guide-to-reprhducible-chde.pdf) British Ecological Society

# TODO

- Load data +
- Reorganise +
- Add solar azimuth +
- Export +
- Adjust times for dome videos +
- Incorporate calibrated dances +
- Remove duplicates

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load functions and packages

Load the packages and other functions that will be used during reorganisation.

```{r load-functions}
source('HS_functions.R')
```


# Load data


## Load the raw data

Load the raw data exported from Excel to `csv`.

```{r load-the-data}
path_file = file.path("0Data",
                      c(
                        "H_calib.csv",#horizontal and dome calibrated dances 
                        "V_uncalib.csv"#vertical dances (not yet calibrated0)
                        )
)
if(any(!file.exists(path_file)))
{print(paste(path_file, '\nnot found'))}
hd = lapply(X = path_file,
            FUN = read.table,
                header = T, 
                sep  = ',')
```


## Combine calibrated and uncalibrated data
Check that all column names match across files loaded.
```{r check-datasets}
nm_vert = names(hd[[1]])
nm_calib = names(hd[[2]])
if(any(!(nm_calib %in% nm_vert) |
       !(nm_vert %in% nm_calib))
   )
{
  message("horizontal columns not in vertical")
print(nm_calib[!(nm_calib %in% nm_vert)])
  message("vertical columns not in horizontal")
print(nm_vert[!(nm_vert %in% nm_calib)])
}else
{print("columns match in all CSVs")}
```

Bind the tables together one set of rows on top of the other
```{r combine-vertical-horizontal}
hd = do.call(what = rbind,
             args = hd)
```

## Inspect data

Exported CSVs sometimes end up with empty rows at the end. 
Check that the last row contains data.

```{r check-last-row}
print(hd[dim(hd)[1],])
```

Last row contains at least some data, placeholder?

<details>

<summary>Click to expand raw data</summary>

```{r inspect-data}
print(hd)
```



# Reorganise

```{r check-column-names}
names(hd)
```


## Rename variables

Rename some variables with shorter or easier to recognise names. 
Video names currently include the full path for the user that annotated them 
these can be simplified to just file name and folder (date recorded).
There are two kinds of grouping factor, firstly the identity of the individual bee,
simplified here to `ID`, and within each individual bee their individual dances,
specified here as `ID_d`.
The date is now in `YYYYDDMM` format, this will be reformatted later.


```{r rename-variables}
#from the list above, identify unused or renamed columns to remove
rm_lst = c(
  "recording_type", #not used currently
  "source_batch", #not used currently
  "dance_key", #not used currently
  "dance_id", #renamed
  "bee_id", # renamed
  "date", #renamed
  "time_dance", #renamed
  # [8,] "waggle_run"
  # [9,] "condition"
  "orientation_method", #not currently used
  # [11,] "csv_file"
  "video_name", #should be identifiable from video_file
  # [13,] "direction_u"
  # [14,] "direction_v"
  # [15,] "video_angle_deg"
  # [16,] "true_north_video_deg"
  # [17,] "waggle_true_bearing_deg"
  # [18,] "mean_magnetic_north_deg"
  # [19,] "true_north_calibration_deg"
  "compass_circular_sd_deg", #not currently used
  "calibration_n_annotations", #not currently used
  # [22,] "calibration_quality_flag"
  "reference_compass_file", #not currently used
  "assigned_calibration_file", #not currently used
  "calibration_id", #not currently used
  "calibration_file", #not currently used
  "reference_landmark_file", #not currently used
  "landmark_dance_file", #not currently used
  "landmark_calibration_deg", #not currently used
  "landmark_calibration_circular_sd_deg", #not currently used
  "north_relative_to_landmark_deg", #not currently used
  "landmark_dance_video_deg", #not currently used
  "landmark_dance_circular_sd_deg", #not currently used
  # [34,] "start_x"
  # [35,] "start_y"
  "start_xy", #same as start_x, start_y
  "start_frame", #not currently used
  "notes"
  # [39,] "calibration_R"
  # [40,] "declination_deg"
)
hd = within(hd, {
  ID = paste0(bee_id) # individual identifier
  ID_d = paste0(dance_id, "_", time_dance, "_", date) # dance identifier
  time = time_dance
  day = date # this will need to be converted to POSIX format

  video_file = file.path(
    basename(dirname(video_name)), #just the folder
    basename(video_name)
  ) # and file name
  #clean up information that has been doubled or will not be used
  rm(list = rm_lst)
})
#confirm resulting columns
names(hd)
```



## Identify factor levels
Some of the character variables could be treated as factors. 
Inspect the different labels to each to check these are informative.


<details>

<summary>Click to expand full list</summary>

```{r summarise-factors}
nm_fact = c(
  # [1,] "waggle_run"
  "condition",
  # "csv_file", #easier to keep file paths as characters
  # [4,] "direction_u"
  # [5,] "direction_v"
  # [6,] "video_angle_deg"
  # [7,] "true_north_video_deg"
  # [8,] "waggle_true_bearing_deg"
  # [9,] "mean_magnetic_north_deg"
  # [10,] "true_north_calibration_deg"
  "calibration_quality_flag",
  # [12,] "start_x"
  # [13,] "start_y"
  # [14,] "calibration_R"
  # [15,] "declination_deg"
  # "video_file", #easier to keep file paths as characters
  # [17,] "day"
  # [18,] "time"
  "ID_d",
  "ID"
)
  
sapply(X = hd[, nm_fact],
       unique
     )
```

## Reformat variables

Reformat variables to factor and numeric. Everything else remains character format by default.


```{r reformat-variables}
nm_num = c(
  "waggle_run",
  # "condition",
  # "csv_file",
  "direction_u",
  "direction_v",
  "video_angle_deg",
  "true_north_video_deg",
  "waggle_true_bearing_deg",
  "mean_magnetic_north_deg",
  "true_north_calibration_deg",
  # "calibration_quality_flag",
  "start_x",
  "start_y",
  "calibration_R",
  "declination_deg",
  # "video_file",
  "day"
  # "time" # this is character but neither factor nor numeric
  # "ID_d",
  # "ID"
)
#format factors
for(nm in nm_fact)
{
hd[, nm] = as.factor(hd[, nm])
}

#format numeric
for(nm in nm_num)
{
hd[, nm] = as.numeric(hd[, nm])
}

summary(hd)
```

There appears to have been an autofill error in condition labelling. 
Rename all unexpected stimuli to `deg_10`, 10° above horizontal.
*20260818* Has this been fixed? Could skip.
*20260825* Currently no `deg_10` data, will keep in case this is added later.

```{r reformat-condition}
hd = within(hd,
            {
              condition = sapply(as.character(condition),
                                switch,
                                vertical = "vertical",
                                horizontal = "horizontal",
                                antisolar = "antisolar",
                                solar = "solar",
                                zenith = "zenith",
                                deg_10 = "deg_10",
                                "deg_10")#otherwise always deg_10
            }
            )
#is there any data for deg_10?
with(subset(hd, condition %in% "deg-10"),
     summary(video_angle_deg))

with(hd, unique(condition))
```


## Check variable structure

Each dance should have only one time code, one condition and one bee.

```{r check-structure}
all(
#individuals per dance
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {length(unique(ID))} )
               }
  )
) == 1,
#times per dance
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {length(unique(time))} )
               }
  )
) == 1,
#days per dance
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {length(unique(time))} )
               }
  )
) == 1,
#stimuli per dance
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {length(unique(condition))} )
               }
  )
) == 1
)

#how many waggle runs per dance?
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {max(waggle_run)} )
               }
  )
)
```

# Add sun location
We can now use the time and date for each dance, combined with the location of 
the hive and feeder, to work out the relative positions of the hive, sun and feeder. 
This tells us what direction each dance should indicate for the feeder, as well as 
any potential effects of sun elevation on accuracy or skylight quality.

## Check times and dates

```{r check-time-date}
#firstly, remove any dances that include neither video nor data
empty_dances = with(subset(hd,
                           subset = is.na(video_angle_deg) &
                                     video_file %in% "/"), # result of file.path('','')
                    unique(ID_d)
                    )
if(empty_dances)
{
  print("removing empty dances:")
  print(as.character(t(t(empty_dances))))
hd = subset(hd,
            subset = !(ID_d %in% empty_dances)
            )
}

missing_time = c('na', '')

#how many dances have day missing
with(subset(hd, is.na(day)),
     length(unique(ID_d))
     )
#very few, which ones?
print(
with(subset(hd, is.na(day)),
     t(t(unique(video_file)))
     )
)
#some of these can be determined from the video name
hd = within(hd,
            {
              #guess the date
              day_guess = sub(x = dirname(video_file),#just the folder
                                                pattern = "_new$", #remove "_new"
                                               replacement = "")
            }
            )
#no guesses are NA
with(subset(hd, is.na(day_guess)),
     length(unique(ID_d))
     )
#one guess is not a date
with(subset(hd, !(as.numeric(day_guess) %in% as.numeric(day[!is.na(day)]))),
     unique(day_guess)
     )

#replace NAs with guesses
hd = within(hd,
            {
            day = ifelse(test = is.na(day) & #if day is NA
                           as.numeric(day_guess) %in% #guess is one of the
                           as.numeric(day[!is.na(day)]),#known dates
                          yes = day_guess,
                          no = day) # don't replace unless sure
            rm('day_guess')#no longer needed
            }
            )
#check after updating
with(subset(hd, is.na(day)),
     length(unique(ID_d))
     )

#how many have time missing
with(subset(hd, time %in% missing_time),
     length(unique(ID_d))
     )
#how many have either
with(subset(hd, is.na(day) | time %in% missing_time),
          length(unique(ID_d))
     )
#how many cannot be converted to calendar dates
hd = within(hd,
       {
       date_fail = is.na(
              as.POSIXct(mapply(FUN = paste, 
                                 day,
                                 time), 
                                      format = "%Y%m%d %H:%M:%OS")#time in YYYY-MM-DD HH:MM:SS
)
              }
              )
with(subset(hd, date_fail),
     length(unique(ID_d))
     )

#Of these how many are missing waggle runs?
# print(with(subset(hd, subset = date_fail), mean(is.na(video_angle_deg)) ), digits = 2)

#which videos are missing time or date
print(
  with(subset(x = hd,
            subset =  is.na(day) | time %in% missing_time),
     {t(t(as.character(unique(video_file))))})
)

#which videos cannot be converted but are not missing time
print( 
  with(subset(x = hd,
            subset =  date_fail & !(time %in% missing_time) ),
     {
       t(t(as.character(unique(video_file))))
       }
     )
 )


```
There is one empty dance that can be removed (possibly extra empty rows at end of CSV).
A total of 4 dances have incomplete time or date, all of which are missing date currently.
These seem to correspond to 4 videos, for which the date can be guessed from the folder name.
It will now be possible to calculate sun azimuth for all remaining dances.



## Set up times

First we need to combine dates in `YYYYMMDD` with times in `HH:MM:SS` to produce a 
 POSIX/C99 "calendar time" for each. This is then converted to UTC for sun location estimation.

```{r add-time-code}

hd = within(hd,
            {
              time_code = mapply(FUN = paste, 
                                 day,
                                 time)#time in YYYY-MM-DD HH:MM:SS
              sast_time =  as.POSIXct(time_code, 
                                      tz = "Africa/Johannesburg", 
                                      format = "%Y%m%d %H:%M:%OS")#South African Standard time
                                      # format = "%Y-%m-%d %H:%M:%OS")#South African Standard time
              utc_time =  as.POSIXct(sast_time, tz = "UTC")#UTC time
              rm("time_code")#now duplicated
            }
)
with(hd, summary(utc_time))
```

In all cases time or date were available and could be converted, so it will 
be possible to calculate the sun's azimuth for all of these.

### Adjust for laptop clock (check this)

Videos recorded on the laptop have time stamps in CET not SAST. This means that 
they are one hour behind, since South African Summer time is `UCT + 2`, while 
Central European Winter is `UTC + 1`. 
Adjust them by adding one hour (60 minutes x 60 seconds) before calculating sun position.

```{r adjust-CET-SAST}
hd = within(hd,
            {
              sast_time = as.POSIXct(
                mapply(FUN = switch,
                                 EXPR = condition,
                                 vertical = sast_time, # camcorder
                                 horizontal = sast_time, # camcorder
                                 deg_10 = sast_time,  # camcorder
                                 antisolar = sast_time + 60*60, #60 min x 60 seconds
                                 solar = sast_time + 60*60,
                                 zenith = sast_time + 60*60,
                                 sast_time
                                 ),
                tz = "Africa/Johannesburg", 
                format = "%Y-%m-%d %H:%M:%OS")#South African Standard time
              #recalculate UTC time
              utc_time =  as.POSIXct(sast_time, tz = "UTC")#UTC time
            }
            )


```




## Inspect experiment times

Convert these time codes to just time-of-day and plot a histogram to check that 
they fit with recorded times of experiments.

```{r inspect-times}
with(hd,
     hist(x = GetToD(sast_time,
                     tz = 'UTC'),
          breaks = 12*60/15 +1,
          xlim = c(7, 19)*60*60,
          freq = TRUE,
          main = "Recorded dances by time",
          xlab = "Local time",
          axes = FALSE,
          border = NA),
     )
axis(side = 1,
     at = 7:19*60*60,
     labels = 7:19)
axis(side = 2)
#add lines indicating the approximate expected start and end of each experiment
abline(v = c(morning_start = 7.50*60*60,
            morning_end = 9.00*60*60,
            midday_start = 11.00*60*60,
            midday_end = 13.00*60*60,
            afternoon_start = 15.00*60*60,
            afternoon_end = 16.50*60*60,
            evening_start = 17.50*60*60,
            evening_end = 19.00*60*60
             ),
       col = c(morning_start = 'darkred',
            morning_end = 'darkred',
            midday_start = 'orange',
            midday_end = 'orange',
            afternoon_start = 'darkred',
            afternoon_end = 'darkred',
            evening_start = 'seagreen',
            evening_end = 'seagreen'
             ),
       lty = c(1,2),
       lwd = 3
       )
```

Most dances appear during the target windows, with the exception of the late afternoon (16:30–17:30). 
It is possible that during periods of high activity we continued to record dances between 
the afternoon and evening experiments, at the transition from medium to low elevations.

## Estimate sun location

In order to determine the sun's location (azimuth and elevation) we need to combine the time 
with the geographic location. `GetSaz` is a vectorised wrapper for `oce::sunAngle`, which corrects 
for refraction, but not the altitude from which the sun is viewed.

The current estimate of hive location is from a combination of Google Maps bookmarks 
and photograph metadata, cross-references with Google Earth satellite images.

```{r add-sun}
# Hive position
# 25° 34' 18" S, 31° 10' 48" E
#-25.57167,      31.18000 

hive = r"(25° 34' 18" S, 31° 10' 48" E)" # N.B. this string contains literal " so needs wrapping in r"()"

# Define coordinates (lon, lat)
hive = Dms2decimal(hive) 

#add sun azimuth
hd = within(hd, 
            {
            sun_az = mapply(
                   FUN = GetSaz,
                   tm = utc_time,
                   lon = hive['longitude'],
                   lat = hive['latitude']
                   )
            }
            )

#add sun azimuth
hd = within(hd, 
            {
            sun_el = mapply(
                   FUN = GetSel,
                   tm = utc_time,
                   lon = hive['longitude'],
                   lat = hive['latitude']
                   )
            }
            )

print(with(hd, sapply(X = list(sun_az, sun_el),FUN =  summary)),
      digits = 2)
```

At the extremes we see Northerly azimuths (015°,340°), presumably around  midday,
while the most common sun azimuths are Westerly ($\approx$270°) in the afternoon and evening. 
Solar elevations range between -8° (after sunset) and 84° (high noon). Around 25% 
are at high elevations (>75°).


### Add Feeder angle

To determine the expected dance directions, we need to know the azimuth angle between 
the sun and the feeder, for which we need the bearing from the hive to the feeder. 
As for the hive position, the current estimate of feeder position was determined  
from a combination of Google Maps bookmarks and photograph metadata, 
cross-referenced with Google Earth satellite images.

`geosphere::bearing` requires coordinates in decimal, with longitude first and then latitude,
so `Dms2decimal` converts from degrees, minutes and seconds to decimal and reorders 
the (latitude,longitude) vector.

```{r add-feeder-angle}
# Hive position
# 25° 34' 18" S, 31° 10' 48" E
#-25.57167,      31.18000 

# Feeder position
# 25° 34' 14" S, 31° 10' 59" E
#-25.57056,      31.18306

hive = r"(25° 34' 18" S, 31° 10' 48" E)" # N.B. this string contains literal " so needs wrapping in r"()"

feeder = r"(25° 34' 14" S, 31° 10' 59" E)"

# Define coordinates (lon, lat)
hive = Dms2decimal(hive) 
feeder =  Dms2decimal(feeder)
#N.B. Dms2decimal converts from (latitude, longitude) to (longitude, latitude)
# to conform with geosphere::bearing

# Calculate bearing from hive to feeder
feeder_bearing = bearing(hive, feeder)
print(feeder_bearing, digits = 3)
feeder_distance = distGeo(hive, feeder)
print(feeder_distance, digits = 3)

#add feeder azimuth relative to sun position
hd = within(hd, 
            {
            feeder_az = Mod360.180(feeder_bearing - sun_az)
            }
            )

print(with(hd, summary(feeder_az)), digits = 3)
print(with(hd, summary(abs(feeder_az))), digits = 3)
```

Based on geographic position alone, we can conclude that the feeder was at 
a bearing of 068° (approx. East-North-East) from the hive, $\approx$ 330m away. 
Given the hilly terrain, the ground distance flown may have been slightly further. 
After subtracting the sun's azimuth, we see that the sun ranged between 
nearly 180° West of the feeder (presumably sunset) and precisely aligned with the 
feeder (presumably during the morning).


### Plot sanity checks

Plot these to get an idea of how they match up.

```{r sanity-plot}

par(mfrow = c(1,2))
with(subset(hd, condition %in% c("vertical", "horizontal", "deg10")),
     {
     plot(x = GetToD(sast_time,
                     tz = 'UTC'),
          y = sun_el,
          xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
          ylim = c(-1,1)*c(5,90),
          xlab = 'UTC',
          ylab = 'Sun Elevation',
          main = 'Camcorder Videos',
          pch = 21,
          col = 'gray50',
           bg = adjustcolor(sapply(X = condition,
                        FUN = switch,
                        vertical = 'darkred',
                        horizontal = 'darkgreen',
                        deg10 = 'seagreen',
                        'seagreen'),
                        alpha.f = 0.25)
     )
     abline(h = 0:6*c(15),
            lty = 2) #limits
     abline(v =  12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
            lty = 3)
     }
     )
with(subset(hd, condition %in% c("solar", "antisolar")),
     {
     plot(x = GetToD(sast_time,
                     tz = 'UTC'),
          y = sun_el,
          xlim = c(20400,78060),#"06:40:00 CET" "22:41:00 CET"
          ylim = c(-1,1)*c(5,90),
          xlab = 'UTC',
          ylab = 'Sun Elevation',
          main = 'USB Camera Videos',
          pch = 21,
          col = 'gray50',
           bg = adjustcolor(sapply(X = condition,
                        FUN = switch,
                        solar = 'blue4',
                        antisolar = 'cyan4',
                        'seagreen'),
                        alpha.f = 0.25 )
          )
     abline(h = 0:6*c(15),
            lty = 2) #limits
     abline(v = 12*60*60, #as.numeric(as.POSIXct("1970-01-01 12:00:00 UTC"))
            lty = 3)
     }
     )

```

# Format angles

```{r investigate-angles}
#check angle format

est_err = with(hd, {
    (atan2(y = direction_v, x = direction_u) + pi/2) - rad(video_angle_deg) 
  })
abs_est_err = mod_circular(abs(est_err))
#these can be out by as much as 1/100 of a degree, except two which are out by 90°
hist(x = deg(abs_est_err), 
     breaks = 1e2,
     xlab = 'absolute error in video angle')
cbind(with(hd[abs_est_err > 1,],
     as.character(unique(ID_d))),
          unique(deg(abs_est_err[abs_est_err > 1])) )
with(hd[abs_est_err > 1,],
           cbind(
     as.character(unique(ID_d)),
          unique(paste(direction_v, direction_u)), 
     unique(as.character(video_file))
     )
)
#these are zero length vectors
#set them to NA
hd = within(hd, 
            {
              direction_v[direction_v ==0 &
                          direction_u ==0
                          ] = NA
              direction_u[direction_v ==0 &
                          direction_u ==0
                          ] = NA
              video_angle_deg[direction_v ==0 &
                          direction_u ==0
                          ] = NA
            }
            )
```
`waggle_true_bearing` should be the direction relative to North.
This has not been calculated yet for this data.
There is a true North for the horizontal condition

```{r check-calibration}
with(subset(hd, condition %in% "horizontal"),
     {
       print(
     summary(true_north_video_deg) 
       )
       print(
     summary(declination_deg) 
       )
       print(
         summary(waggle_true_bearing_deg),
         digits = 3
       )
       print(
         summary(waggle_true_bearing_deg - video_angle_deg),
         digits = 3
       )       
       
       print(
         summary(true_north_video_deg),
         digits = 3
       )
       #check the correction
       print(
         summary(Mod360.180(video_angle_deg - true_north_video_deg - waggle_true_bearing_deg)),
         digits = 3
       )
       
     }
     )

```



```{r format-angles}
#assign angles to numeric

#video_angle_deg is the dance angle clockwise rel to vertical, except  error
hd = within(hd, 
            {
              # bearing = deg(atan2(y = direction_v, x = direction_u) + pi/2) %% 360
              bearing = mapply(FUN = switch,
                               EXPR = condition,
                               horizontal = waggle_true_bearing_deg, #corrected with compass
                               deg10 = waggle_true_bearing_deg,
                               vertical =  video_angle_deg, #aligned by default
                               # antisolar = video_angle_deg, #aligned by default
                               # solar = (video_angle_deg +180) %% 360, #misaligned 180°
                               antisolar = waggle_true_bearing_deg, #now calibrated
                               solar = waggle_true_bearing_deg, #now calibrated
                               zenith = waggle_true_bearing_deg, #now calibrated
                               video_angle_deg
                              )
              signed_angle = Mod360.180(bearing)  # bearing between -180 and 180
              # angle = circular(rad(signed_angle),# bearing between -pi and pi
              #                  rotation = 'clock') # circular format suppresses later warnings
              angle = rad(signed_angle)# circular format is not needed for organising
                               
              rm('video_angle_deg')
            }
            )
with(hd, sapply(list(bearing, signed_angle, angle), FUN = summary))
```
# Inspect dances

### Plot an example dance

For example we can look at the dances for Blue-22 recorded on 2025-01-28 at 
07:57:05, 08:10:00, 08:29:00, 08:34:00 & 10:44:09—each of which is the first for 
one of the conditions.
Sun azimuth (possibly incorrect) is shown as an orange line.

```{r example-subset}
subs_20250128 = subset(hd, day %in% "20250128")
#find first dance for each condition on 2025.01.28
example_lst = list()
for(stim in unique(hd$condition))
{
example_lst[[stim]] = with(subset(subs_20250128, condition %in% stim),
                           unique(ID_d)[1])
}
#inspect
print(t(t(unlist(example_lst))))

example_subs = subset(x = hd,
                      ID_d %in% paste(unlist(example_lst))
                      )

par(mfrow = c(2,2), mar = c(0,0,0,0))
with(subset(example_subs,
            subset = condition %in% 'vertical'),
     {
  PCfun(angles = bearing,
        col = 'darkgreen',
        shrink = 1.5,
        title = paste(unique(time), unique(condition)))
  lines(
    x = c(0, sin(rad(feeder_az))),
    y = c(0, cos(rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)

with(subset(example_subs,
            subset = condition %in% 'horizontal'),
     {
PCfun(angles = bearing,
      col = 'darkred',
      shrink = 1.5,
      title = paste(unique(time), unique(condition)))
         lines(
    x = c(0, sin(rad(feeder_az))),
    y = c(0, cos(rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
with(subset(example_subs,
            subset = condition %in% 'antisolar'),
     {
PCfun(angles = bearing,
      col = 'cyan4',
      shrink = 1.5,
      title = paste(unique(time), unique(condition)))
         lines(
    x = c(0, sin(rad(feeder_az))),
    y = c(0, cos(rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
with(subset(example_subs,
            subset = condition %in% 'solar'),
     {
PCfun(angles = bearing,
      col = 'blue4',
      shrink = 1.5,
      title = paste(unique(time), unique(condition)))
         lines(
    x = c(0, sin(rad(feeder_az+180))),
    y = c(0, cos(rad(feeder_az+180))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
```



# Calculate dance offset from feeder
```{r feeder-angle}
#assign angles to numeric

#video_angle_deg is the dance angle clockwise rel to vertical, except  error
hd = within(hd, 
            {
              feeder_angle = mod_circular(rad(bearing - feeder_az))# angle between dance and expected dance
              abs_feeder_angle_deg = deg(abs(as.numeric(feeder_angle)))# absolute angle in numeric format
            }
            )

print(
  with(subset(hd, condition %in% 'vertical'), summary(abs_feeder_angle_deg)),
  digits = 2
)
print(
  with(subset(hd, condition %in% 'horizontal'), summary(abs_feeder_angle_deg)),
  digits = 1
)
print(
  with(subset(hd, condition %in% 'antisolar'), summary(abs_feeder_angle_deg)),
  digits = 1
)
print(
  with(subset(hd, condition %in% 'solar'), summary(abs_feeder_angle_deg)),
  digits = 1
)

```
### Plot example feeder angles

```{r feeder-angle-examples}
#update the subset
example_subs = subset(x = hd,
                      ID_d %in% paste(unlist(example_lst))
                      )

par(mfrow = c(2,2), mar = c(0,0,0,0))
    with(subset(example_subs,
                subset = condition %in% 'vertical'),
         {
      PCfun(angles = deg(feeder_angle),
            col = 'darkgreen',
            shrink = 1.5,
            title = paste(unique(time), unique(condition)))
         }
    )

with(subset(example_subs,
            subset = condition %in% 'horizontal'),
     {
PCfun(angles = deg(feeder_angle),
      col = 'darkred',
      shrink = 1.5,
      title = paste(unique(time), unique(condition)))
     }
)
with(subset(example_subs,
            subset = condition %in% 'antisolar'),
     {
PCfun(angles = deg(feeder_angle),
      col = 'cyan4',
      shrink = 1.5,
      title = paste(unique(time), unique(condition)))
     }
)
with(subset(example_subs,
            subset = condition %in% 'solar'),
     {
PCfun(angles = deg(feeder_angle),
      col = 'blue4',
      shrink = 1.5,
      title = paste(unique(time), unique(condition)))
     }
)
```



# Write out reorganised data

```{r write-data}
# Write out table

write.table(x = hd,
            file = file.path(unique(dirname(path_file))[1], "hs_dance_reorg.csv"),
            sep = ',',
            row.names = FALSE
            )
```
