---
title: "Organise Data"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    fig_caption: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
  # bookdown::pdf_document2:
  #   highlight: rstudio
---

# References

> Natalie Cooper & Pen-Yuan Hsing, 2017 [A Guide to Reprhducible Chde in Ecology and Evolution](https://britishecologicalsociety.org/wp-content/uploads/2017/12/guide-to-reprhducible-chde.pdf) British Ecological Society

# TODO

- Load data +
- Reorganise +
- Add solar azimuth +
- Export
- Incorporate calibration

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load functions and packages

Load the packages and other functions that will be used during reorganisation.

```{r load-functions}
source('HS_functions.R')
```

# Load data

## Load the raw data

Load the raw data exported from Excel to `csv`.

```{r load-the-data}
path_file = "0Data/B22_dances_with_calibrations.csv"
if(!file.exists(path_file))
{print(path_file, '\nnot found')}
hd = read.table(file = path_file, 
                header = T, 
                sep  = ',')
```

## Inspect data

Exported CSVs sometimes end up with empty rows at the end. 
Check that the last row contains data.

```{r check-last-row}
print(hd[dim(hd)[1],])
```

<details>

<summary>Click to expand raw data</summary>

```{r inspect-data}
print(hd)
```

# Reorganise

```{r check-column-names}
names(hd)
```
## Rename variables

Rename some variables with shorter or easier to recognise names. 
Video names currently include the full path for the user that annotated them 
these can be simplified to just file name and folder (date recorded).


```{r rename-variables}
hd = within(hd, {
  ID = paste0(dancer_id) # individual identifier
  ID_d = paste0(dance_id, "_", 
                time, "_",
                source_batch) # dance identifier
  day = source_batch # this is already in POSIX format
  video_file = file.path(basename(dirname(video_name)), #just the folder
                         basename(video_name)) # and file name
  #clean up information that has been doubled
  rm(list = c('dancer_id', 'dance_id', 'dance_key', 'source_batch', "date",
              "video_name", #should be identifiable from video_file
              "waggle_directions",#split into direction_u  and direction_v
              "start_xy"#split into "start_x" and "start_y" 
              ))
})
```

## Identify factor levels
Some of the character variables could be treated as factors. 
Inspect the different labels to each to check these are informative.

```{r summarise-factors}
nm_fact = c( "ID",         
 "ID_d"          ,
 # "dance_key"         ,
 # [4,] "waggle_run"        
 "stimulus"         , 
 # "notes"            , 
 # [7,] "waggle_directions" 
#  [8,] "start_x"           
#  [9,] "start_y"           
# [10,] "start_xy"          
# [11,] "csv_file"          
# [12,] "time"              
# [13,] "date"              
# [14,] "direction_u"       
# [15,] "direction_v"       
# [16,] "video_angle_deg"   
# [17,] "start_frame"       
 # "video_name"        
 "video_file" # avoid printing full path        
# [19,] "annotation_session"
# [20,] "source_batch"
            )
sapply(X = hd[, nm_fact],
       unique
     )
```

## Reformat variables

Reformat variables to factor and numeric. Everything else remains character format by default.


```{r reformat-variables}
nm_num = c(
 #  "dancer_id",         
 # "dance_id"          ,
 # "dance_key"         ,
 "waggle_run",        
 # "stimulus"         , 
 # "notes"            , 
 # [7,] "waggle_directions" 
"start_x"    ,       
"start_y"     ,      
# "start_xy"          
# [11,] "csv_file"          
# [12,] "time"              
# [13,] "date"              
"direction_u",
"direction_v",
"video_angle_deg",
"start_frame",
 # "video_name"        
"annotation_session"
# [20,] "source_batch"
)
#format factors
for(nm in nm_fact)
{
hd[, nm] = as.factor(hd[, nm])
}

#format numeric
for(nm in nm_num)
{
hd[, nm] = as.numeric(hd[, nm])
}

summary(hd)
```

There appears to have been an autofill error in stimulus labelling. 
Rename all unexpected stimuli to `deg_10`, 10° above horizontal.
```{r reformat-stimulus}
hd = within(hd,
            {
              stimulus = sapply(as.character(stimulus),
                                switch,
                                vertical = "vertical",
                                horizontal = "horizontal",
                                antisolar = "antisolar",
                                solar = "solar",
                                deg_10 = "deg_10",
                                "deg_10")#otherwise always deg_10
            }
            )
with(hd, unique(stimulus))
```

## Check variable structure

Each dance should have only one time code, one stimulus and one bee.

```{r check-structure}
all(
#individuals per dance
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {length(unique(ID))} )
               }
  )
) == 1,
#times per dance
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {length(unique(time))} )
               }
  )
) == 1,
#days per dance
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {length(unique(time))} )
               }
  )
) == 1,
#stimuli per dance
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {length(unique(stimulus))} )
               }
  )
) == 1
)

#how many waggle runs per dance?
summary(
  sapply(X = unique(hd$ID_d),
               FUN = function(id){
                 with(subset(hd, ID_d %in% id),
                          {max(waggle_run)} )
               }
  )
)
```

## Add sun location

```{r add-time-code}

hd = within(hd,
            {
              time_code = mapply(FUN = paste, 
                                 day,
                                 time)#time in YYYY-MM-DD HH:MM:SS
              sast_time =  as.POSIXct(time_code, 
                                      tz = "Africa/Johannesburg", 
                                      format = "%Y-%m-%d %H:%M:%OS")#South African Standard time
              utc_time =  as.POSIXct(sast_time, tz = "UTC")#UTC time
              rm("time_code")#now duplicated
            }
)
```

```{r add-sun}
# Hive position
# 25° 34' 19" S, 31° 10' 53" E
#-25.571944, 31.181389


#add sun azimuth
hd = within(hd, 
            {
            sun_az = mapply(
                   FUN = GetSaz,
                   tm = utc_time,
                   lon = 31.181389,
                   lat = -25.571944)
            }
            )

#add sun azimuth
hd = within(hd, 
            {
            sun_el = mapply(
                   FUN = GetSel,
                   tm = utc_time,
                   lon = 31.181389,
                   lat = -25.571944)
            }
            )

print(with(hd, sapply(X = list(sun_az, sun_el),FUN =  summary)),
      digits = 2)
```
```{r add-feeder-angle}
# Hive position
# 25° 34' 19" S, 31° 10' 53" E
#-25.571944, 31.181389

# Feeder position
# 25° 34' 14" S, 31° 10' 59" E
#-25.570556, 31.183056

# Define coordinates (lon, lat)
hive = c(longitude = 31.181389, 
         latitude = -25.571944)
feeder =  c(longitude = 31.183056, 
            latitude = -25.570556)

# Calculate bearing from hive to feeder
feeder_bearing = bearing(hive, feeder)
print(feeder_bearing, digits = 3)

#add feeder azimuth
hd = within(hd, 
            {
            feeder_az = Mod360.180(sun_az - feeder_bearing)
            }
            )
print(with(hd, summary(feeder_az)), digits = 3)
```

## Format angles

```{r investigate-angles}
#check angle format

est_err = with(hd, {
    (atan2(y = direction_v, x = direction_u) + pi/2) - rad(video_angle_deg) 
  })
abs_est_err = mod_circular(abs(est_err))
#these can be out by as much as 1/100 of a degree, except two which are out by 90°
hist(x = deg(abs_est_err), 
     breaks = 1e2,
     xlab = 'absolute error in video angle')
cbind(with(hd[abs_est_err > 1,],
     as.character(unique(ID_d))),
          unique(deg(abs_est_err[abs_est_err > 1])) )
with(hd[abs_est_err > 1,],
           cbind(
     as.character(unique(ID_d)),
          unique(paste(direction_v, direction_u)), 
     unique(as.character(video_file))
     )
)
#these are zero length vectors
#set them to NA
hd = within(hd, 
            {
              direction_v[direction_v ==0 &
                          direction_u ==0
                          ] = NA
              direction_u[direction_v ==0 &
                          direction_u ==0
                          ] = NA
              video_angle_deg[direction_v ==0 &
                          direction_u ==0
                          ] = NA
            }
            )
```

```{r format-angles}
#assign angles to numeric

#video_angle_deg is the dance angle clockwise rel to vertical, except  error
hd = within(hd, 
            {
              # bearing = deg(atan2(y = direction_v, x = direction_u) + pi/2) %% 360
              bearing = video_angle_deg
              signed_angle = Mod360.180(video_angle_deg)  # bearing between -180 and 180
              angle = circular(rad(signed_angle),# bearing between -pi and pi
                               rotation = 'clock') # circular format suppresses later warnings
              rm('video_angle_deg')
            }
            )
with(hd, sapply(list(bearing, signed_angle, angle), FUN = summary))
```
# Inspect dances

### Plot an example dance

For example we can look at the dances for Blue-22 recorded on 2025-01-28 at 
07:57:05, 08:10:00, 08:29:00, 08:34:00 & 10:44:09—each of which is the first for 
one of the stimulus conditions.
Sun azimuth (possibly incorrect) is shown as an orange line.

```{r example-subset}
subs_20250128 = subset(hd, day %in% "2025-01-28")
#find first dance for each condition on 2025.01.28
example_lst = list()
for(stim in unique(hd$stimulus))
{
example_lst[[stim]] = with(subset(subs_20250128, stimulus %in% stim),
                           unique(ID_d)[1])
}
#inspect
print(t(t(unlist(example_lst))))

example_subs = subset(x = hd,
                      ID_d %in% paste(unlist(example_lst))
                      )

par(mfrow = c(2,2), mar = c(0,0,0,0))
with(subset(example_subs,
            subset = stimulus %in% 'vertical'),
     {
  PCfun(angles = bearing,
        col = 'darkgreen',
        shrink = 1.5,
        title = paste(unique(time), unique(stimulus)))
  lines(
    x = c(0, sin(-rad(feeder_az))),
    y = c(0, cos(-rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)

with(subset(example_subs,
            subset = stimulus %in% 'horizontal'),
     {
PCfun(angles = bearing,
      col = 'darkred',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus)))
         lines(
    x = c(0, sin(-rad(feeder_az))),
    y = c(0, cos(-rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
with(subset(example_subs,
            subset = stimulus %in% 'antisolar'),
     {
PCfun(angles = bearing,
      col = 'cyan4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus)))
         lines(
    x = c(0, sin(-rad(feeder_az))),
    y = c(0, cos(-rad(feeder_az))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
with(subset(example_subs,
            subset = stimulus %in% 'solar'),
     {
PCfun(angles = bearing,
      col = 'blue4',
      shrink = 1.5,
      title = paste(unique(time), unique(stimulus)))
         lines(
    x = c(0, sin(-rad(feeder_az+180))),
    y = c(0, cos(-rad(feeder_az+180))),
    col = adjustcolor('orange', alpha.f = 0.5),
    lwd = 2
  )
     }
)
```
# Write out reorganised data

```{r write-data}
# Write out table

write.table(x = hd,
            file = file.path(dirname(path_file), "hs_dance_reorg.csv"),
            sep = ',',
            row.names = FALSE
            )
```
